---
title: "changelog"
author: 

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

<hr>

### Note for instructors and TAs

This page records substantive changes to the **UBC edition** of the BIOL
300 R labs. The purpose is to help BIOL 300 instructors and TAs identify any
meaningful changes to the tutorials. The date in the footer reflects
the most recent *substantive* update. Minor changes such as typo fixes,
formatting adjustments, or changes that do not really affect answers may
not require answer-key updates. Check with Vikram if unsure.

<hr>

# 2026-07-06 {#update-2026-07-06}

## Key changes
- Bump R to version 4.6.1
- Revise Lab 1 student data sheet and stash docx & pdf

## Site materials developed with

- R version 4.6.1 (2026-06-24) -- "Happy Hop"
- RStudio 2026.06.0 Build 242
  - "Blue Plumbago" Release (d1dc28bb, 2026-06-18)
- `rmarkdown` 2.31

## R packages used in labs

```{r}
packageVersion("tidyverse")
packageVersion("car")
packageVersion("binom")
```

<hr>

# 2026-06-30 {#update-2026-06-30}

## Key changes
- Word edits to Lab 3
  - Explicit use of indexing with `sample.int()`

## Site materials developed with

- R version 4.6.0 (2026-04-24) -- "Because it was There"
- RStudio 2026.06.0 Build 242
  - "Blue Plumbago" Release (d1dc28bb, 2026-06-18)
- `rmarkdown` 2.31

## R packages used in labs

```{r}
packageVersion("tidyverse")
packageVersion("car")
packageVersion("binom")
```

<hr>

# 2026-06-26 {#update-2026-06-26}

## Key changes
- Correction of fingerlengths data set
- Re-constitution of ABDLabs zip file
- Corrections to Lab 8 question numbering
- Correction to Lab 2, question 3c: switch Africa to Asia
- Add "bad code line" to Lab 8, q5

## Site materials developed with

- R version 4.6.0 (2026-04-24) -- "Because it was There"
- RStudio 2026.05.1 Build 225, "Golden Wattle" Release (89e0c7da,
  2026-06-07)
- `rmarkdown` 2.31

## R packages used in labs

```{r}
packageVersion("tidyverse")
packageVersion("car")
packageVersion("binom")
```

<hr>

# 2026-06-23 {#update-2026-06-23}

## Key changes
- Establishment of "UBC Edition" of materials
- Re-working of labs 1-4
  - Re-framing labs 1 and 2 so that lab 1 = "R as a calculator" whereas 
  lab 2 = "dealing with tabular data"
  - Re-integration of sampling lab as lab 3
  - Shifting of graphics lab to lab 4
- Integration of Activities into Questions
- Updated version of [ABDLabs.zip](ABDLabs.zip)
  - Additon of fingerlengths data into DataForLabs
  - Revisions of Learning the Tools scripts
- [Troubleshooting](troubleshooting.html) page

## Site materials developed with

- R version 4.6.0 (2026-04-24) -- "Because it was There"
- RStudio 2026.05.1 Build 225, "Golden Wattle" Release (89e0c7da,
  2026-06-07)
- `rmarkdown` 2.31

## R packages used in labs

```{r}
packageVersion("tidyverse")
packageVersion("car")
packageVersion("binom")
```


<br>
