---
title: "changelog"
author: 

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

<hr>

### Note for instructors and TAs

This page records substantive changes to the **UBC edition** of the BIOL
300 R labs. The purpose is to help BIOL 300 instructors and TAs keep
answer keys and lesson plans current. The date in the footer reflects
the most recent *substantive* update. Minor changes such as typo fixes,
formatting adjustments, or changes that do not really affect answers may
not require answer-key updates. Check with Vikram if unsure.

<hr>

# 2026-06-26 {#update-2026-06-26}

## Key changes
- Correction of fingerlengths data set
- Re-constitution of ABDLabs zip file
- Corrections to Lab 8 question numbering
- Correction to Lab 2, question 3c: switch Africa to Asia

## Site materials developed with

- R version 4.6.0 (2026-04-24) -- "Because it was There"
- RStudio 2026.05.1 Build 225, "Golden Wattle" Release (89e0c7da,
  2026-06-07)
- `rmarkdown` 2.31

## R packages used in labs

```{r}
packageVersion("tidyverse")
packageVersion("car")
packageVersion("binom")
```

<hr>

# 2026-06-23 {#update-2026-06-23}

## Key changes
- Establishment of "UBC Edition" of materials
- Re-working of labs 1-4
  - Re-framing labs 1 and 2 so that lab 1 = "R as a calculator" whereas 
  lab 2 = "dealing with tabular data"
  - Re-integration of sampling lab as lab 3
  - Shifting of graphics lab to lab 4
- Integration of Activities into Questions
- Updated version of [ABDLabs.zip](ABDLabs.zip)
  - Additon of fingerlengths data into DataForLabs
  - Revisions of Learning the Tools scripts
- [Troubleshooting](troubleshooting.html) page

## Site materials developed with

- R version 4.6.0 (2026-04-24) -- "Because it was There"
- RStudio 2026.05.1 Build 225, "Golden Wattle" Release (89e0c7da,
  2026-06-07)
- `rmarkdown` 2.31

## R packages used in labs

```{r}
packageVersion("tidyverse")
packageVersion("car")
packageVersion("binom")
```


<br>
