---
title: "R code for example in Chapter 7: <br>Analyzing proportions"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
suppressPackageStartupMessages(library(binom))
suppressPackageStartupMessages(library(ggplot2))
```

***

We used R to analyze all examples in Chapter 7. We put the code here so that you can too.

<br>
Also see our [lab on how to carry out frequency analaysis in R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Frequency_data.html).

<br>

# New methods on this page

These are the core commands that produce the new methods described in Chapter 7. The variable names are plucked from the examples further below.

<ul>
<li>Calculating binomial probabilities:

```{r eval = FALSE}
    dbinom(6, size = 27, prob = 0.25)
```

<li>Binomial test:

```{r eval = FALSE}
    binom.test(10, n = 25, p = 0.061)
```

<li>Agresti-Coull 95% confidence interval for the proportion:

```{r eval = FALSE}
    binom.confint(131, n = 169, method = "ac")
```

<li>Other new methods:
<ul>
<li>Install an R package.
<li>Sampling distribution of a proportion by repeated sampling from a known population.
</ul>
</ul> 

***

<br>

# Load required packages 

We'll use the ```ggplot2``` add on package to draw many plots, and the ```binom``` package to calculate a confidence interval for a proportion using the Agresti-Coull method. They need to be loaded only once per session.

You might need to install the ```binom``` package on your computer if you have not already done so.

```{r eval = FALSE}
install.packages("binom", dependencies = TRUE) # only if not yet installed
```

```{r eval = TRUE}
library(binom)
library(ggplot2)
```

***

<br>

# 7.1 Binomial distribution

***Calculate a binomial probability**, the probability of obtaining $X$ successes in $n$ trials when trials are independent and probability of success $p$ is the same for every trial. 
*

<br>

## Binomial probability

This uses Example 6.4 on mirror image flowers. The probability of getting exactly 6 left-handed flowers when $n = 27$ and $p = 0.25$ is

```{r}
dbinom(6, size = 27, prob = 0.25)
```

<br>

## Table of probabilities

Tabulate the probability distribution of all possible values for the number of left-handed flowers out of 27 (Table 7.1-1).

```{r}
xsuccesses <- 0:27
probx <- dbinom(xsuccesses, size = 27, prob = 0.25) 
probTable <- data.frame(xsuccesses, probx)
probTable
```

<br>

## Histogram of probability distribution

This illustrates the binomial distribution for the number of left-handed flowers when $n = 27$ and $p = 0.25$ (Figure 7.1-1).

<!--
```{r}
barplot(height = probx, names.arg = xsuccesses, space = 0, las = 1,
	col = "firebrick", ylab = "Probability", xlab = "Number of left-handed flowers")
```

or -->

```{r}
ggplot(probTable, aes(x = xsuccesses, y = probx)) + 
    geom_bar(stat = "identity", fill = "firebrick", col = "black") +
    labs(x = "Number of left-handed flowers", y = "Probability") +
    theme_classic()
```

***

<br>

## Sampling distribution of a proportion

Compare sampling distributions for the proportion based on $n = 10$ and $n = 100$ (Figure 7.1-2).

Take a large number of random samples of $n = 10$ from a population having probability of success $p = 0.25$. Convert to proportions by dividing by the sample size. Do the same for the larger sample size $n = 100$. The following commands use 10,000 random samples.

```{r}
successes10 <- rbinom(10000, size = 10, prob = 0.25)
proportion10 <- successes10 / 10
successes100 <- rbinom(10000, size = 100, prob = 0.25)
proportion100 <- successes100 / 100
```

<!--
The ```par(mfrow = c(2,1))``` command sets up a graph window that will plot both graphs arranged in 2 rows and 1 column.

```{r}
par(mfrow = c(2,1))
hist(proportion10, breaks = 10, right = FALSE, xlim = c(0,1),
	xlab = "Sample proportion", col = "firebrick", main = "")
hist(proportion100, breaks = 20, right = FALSE, xlim = c(0,1),
	xlab = "Sample proportion", col = "firebrick", main = "")
par(mfrow = c(1,1))
```
or

-->

```{r}
ggplot(data = data.frame(proportion10), aes(x = proportion10)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 0.1) +
    labs(x = "Sample proportion", y = "Probability") + 
    coord_cartesian(xlim = c(0, 1)) +
    theme_classic()

ggplot(data = data.frame(proportion100), aes(x = proportion100)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 0.01) +
    labs(x = "Sample proportion", y = "Probability") + 
    coord_cartesian(xlim = c(0, 1)) +
    theme_classic()
```

***

<br>

# Example 7.2. Sex and the X

*The **binomial test**, used to test whether spermatogenesis genes in the mouse genome occur with unusual frequency on the X chromosome.*

<br>

## Read and inspect data

Each row in the data file represents a different spermatogenesis gene.

```{r}
mouseGenes <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter07/chap07e2SexAndX.csv"), stringsAsFactors = FALSE)
head(mouseGenes)
```

<br>

## Tabulate number of genes

Tally the number of spermatogenesis genes on the X-chromosome and the number not on the X-chromosome.

```{r}
table(mouseGenes$onX)
```

<br>

## Calculate probabilities

Calculate the binomial probabilities of all possible outcomes under the null hypothesis (Table 7.2-1). Under the binomial distribution with $n = 25$ and $p = 0.061$, the number of successes can be any integer between 0 and 25.

```{r}
xsuccesses <- 0:25
probx <- dbinom(xsuccesses, size = 25, prob = 0.061)
data.frame(xsuccesses, probx)
```

<br>

## Calculate $P$-value

Use these probabilities to calculate the $P$-value corresponding to an observed 10 spermatogenesis genes on the X chromosome. Remember to multiply the probability of 10 or more successes by 2 for the two-tailed test result.

```{r}
2 * sum(probx[xsuccesses >= 10])
```

<br>

## R's binomial test

For a faster result, try R's built-in binomial test. The resulting $P$-value is slightly different from our own calculation. In the book, we get the two-tailed probability by multiplying the one-tailed probability by 2. However, computer programs may calculate the probability of extreme results at the "other" tail with a different method. The output of ```binom.test()``` includes a confidence interval for the proportion using the Clopper-Pearson method, which is more conservative than the Agresti-Coull method.

```{r}
binom.test(10, n = 25, p = 0.061)
```

***

<br>

# Example 7.3. She-turtles

***Standard error and 95% confidence interval for a proportion** using the Agresti-Coull method for the confidence interval.*

<br>

## Read and inspect data

```{r}
turtles <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter07/chap07e3GreenTurtleSex.csv"), stringsAsFactors = FALSE)
head(turtles)
```

<br>

Frequency table of female and male offspring number.

```{r}
turtleTable <- table(turtles$Sex)
turtleTable
```

<br>

## Calculate proportion

Estimate the proportion of offspring that are female.

```{r}
n <- sum(turtleTable)
n
X <- turtleTable[1]
pHat <-  X / n
pHat
```

<br>

## Standard error of proportion

```{r}
sqrt( (pHat * (1 - pHat))/n )
```

<br>

## Agresti-Coull 95% CI

First, calculate the Agresti-Coull 95% confidence interval for the population proportion using the formula in the book.

```{r}
pPrime <- (X + 2)/(n + 4)
pPrime
lower <- pPrime - 1.96 * sqrt( (pPrime * (1 - pPrime))/(n + 4) )
upper <- pPrime + 1.96 * sqrt( (pPrime * (1 - pPrime))/(n + 4) )
c(lower = lower, upper = upper)
```

Second, use the ```binom``` package. The result will be very slightly different from the one you calculated above because the formula we use in the book takes a slight shortcut.

```{r}
binom.confint(X, n = n, method = "ac")
```