---
title: "R code for example in Chapter 12: <br>Comparing two means"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

We used R to analyze all examples in Chapter 12. We put the code here so that you can too.

<br>
Also see our [lab on comparing two means using R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Comparing_means_of_2_groups.html).

<br>

# New methods on this page

<ul>
<li>Paired *t*-test and 95% confidence interval of a mean difference:

```{r eval = FALSE}
t.test(blackbird$logAfterImplant, blackbird$logBeforeImplant, paired = TRUE)
```

<li>Two-sample *t*-test and 95% confidence interval for a difference between two means:

```{r eval = FALSE}
t.test(squamosalHornLength ~ Survival, data = lizard, var.equal = TRUE)
```

<li>Other new methods:
<ul>
<li>A multiple histogram plot with ```ggplot()```.
<li>The *F* test to compare two variances.
<li>The 95% confidence interval for the variance ratio.
<li>Levene's test to compare variances between two (or more) groups.
<li>Convert data in "wide" format to "long" format.
</ul>
</ul>

***

<br>

# Load required packages 

We'll use the ```ggplot2``` add on package to make graphs, and ```dplyr``` to calculate summary statistics by group. The ```car``` package is needed only for the optional Levene's test. You may need to install it first onto you computer if you have not already done so.

```{r eval = FALSE}
install.packages("car", dependencies = TRUE) # install only if necessary
```

```{r message = FALSE}
library(ggplot2)
library(dplyr)
library(car)
```

***

<br>

# Example 12.2. Red-winged blackbirds

***Confidence interval for mean difference**** and the **paired *t*-test**, comparing immunocompetence of red-winged blackbirds before and after testosterone implants.*

<br>

## Read and inspect data

The data are in "wide" format (one row per individual, with the response variable at two time points in separate columns).

```{r eval = TRUE}
blackbird <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter12/chap12e2BlackbirdTestosterone.csv"), stringsAsFactors = FALSE)
blackbird
```

<br>

## Plot paired data

A strip chart for paired data, with lines connecting dots for the same individual (Figure 12.2-1). We first need to put the log-transformed data into "long" format, with the before and after measurements in the same column of data, and a new variable to indicate the Before and After data. 

We did this by making a separate data frame for Before and After treatments (they must having identical variable names) and then ```rbind()``` them. We then ordered the categories of the treatment variable so that Before data comes first in the plot.

```{r eval = TRUE}
beforeData <- data.frame(blackbird = blackbird$blackbird, 
                  logAntibody = blackbird$logBeforeImplant,
                  time = "Before")
afterData <- data.frame(blackbird = blackbird$blackbird, 
                  logAntibody = blackbird$logAfterImplant,
                  time = "After")
blackbird2 <- rbind(beforeData, afterData)
blackbird2$time <- factor(blackbird2$time, levels = c("Before", "After"))
```

A ```ggplot()``` command for a paired data plot. The argument ```alpha =``` plots transparent dots, which helps to distinguish visually any overlapping measurements.

```{r eval = TRUE}
ggplot(blackbird2, aes(x = time, y = logAntibody)) +  
    geom_point(size = 5, col = "firebrick", alpha = 0.5) + 
    geom_line(aes(group = blackbird)) +
    labs(x = "Implant treatment", y = "Antibody production rate (ln[mOD/min])") + 
    theme_classic()
```

<br>

## Paired differences

We make a new variable ```d``` with the After minus Before differences.

```{r eval = TRUE}
blackbird$d <- blackbird$logAfterImplant - blackbird$logBeforeImplant
head(blackbird)
```

<br>

## Histogram of differences

The commands below generate Figure 12.2-2.

```{r eval = TRUE}
hist(blackbird$d, right = FALSE, col = "firebrick", las = 1)
```

or

```{r eval = TRUE}
ggplot(blackbird, aes(x = d)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.1, 
        boundary = 0, closed = "left") + 
    labs(x = "Difference (after–before)", y = "Frequency") + 
    theme_classic()
```

<br>

## 95% CI for mean difference

95% confidence intervals are part of the output of the ```t.test()``` function, viewed in isolation by adding ```$conf.int``` to the command.

```{r eval = TRUE}
t.test(blackbird$d)$conf.int
```

Or, we can use the ```paired = TRUE``` argument of ```t.test()``` and specifying the pair of variables.

```{r eval = TRUE}
t.test(blackbird$logAfterImplant, blackbird$logBeforeImplant, paired = TRUE)$conf.int
```

<br>

## Paired *t*-test

A paired *t*-test can be done either on differences you have already calculated (```d``` here) or by using the ```paired = TRUE``` argument with the measurements from the two groups.

```{r eval = TRUE}
t.test(blackbird$d)
```

or

```{r eval = TRUE}
t.test(blackbird$logAfterImplant, blackbird$logBeforeImplant, paired = TRUE)
```

***

<br>

# Example 12.3. Horned lizards

***Confidence interval for the difference between two means**, and the **two-sample *t*-test**, comparing horn length of live and dead (spiked) horned lizards.*

<br>

## Read and inspect data

```{r eval = TRUE}
lizard <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter12/chap12e3HornedLizards.csv"), stringsAsFactors = FALSE)
head(lizard)
```

Note that there is one missing value for the variable "squamosalHornLength". 

<br>

## Multiple histograms

We begin by ordering the categories of the survival variable so that "living" comes before "killed" in the histogram. The plot command will deliver a warning to let you know that one case was dropped (the one lizard missing a horn length measurement).

```{r eval = TRUE}
lizard$Survival <- factor(lizard$Survival, levels = c("living", "killed"))

ggplot(lizard, aes(x = squamosalHornLength)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 2, 
        boundary = 0, closed = "left") +
    facet_wrap( ~ Survival, ncol = 1, scales = "free_y") +
    labs(x = "Horn length (mm)", y = "Frequency") + 
    theme_classic()
```

<br>

## 95% CI for difference between means

The 95% confidence interval for the difference between two means is part of the output of ```t.test()```. Append ```$confint``` after calling the function to get R to report only the confidence interval.

The formula in the following command tells R to compare ```squamosalHornLength``` between the two groups indicated by ```Survival```. When using the formula method, the data frame containing the variables must be indicated by ```data = ````. The ```var.equal =``` argument indicates that we assume the variances of the two groups to be equal (the default is ```FALSE```).

```{r eval = TRUE}
t.test(squamosalHornLength ~ Survival, data = lizard, var.equal = TRUE)$conf.int
```

<br>

## Two-sample *t*-test

A test of the difference between the two means can be carried out with ```t.test()```. The formula in the following command tells R to compare ```squamosalHornLength``` between the two groups indicated by ```Survival```. When using the formula method, the data frame containing the variables must be indicated by ```data = ````. The ```var.equal =``` argument indicates that we assume the variances of the two groups to be equal (the default is ```FALSE```).

```{r eval = TRUE}
t.test(squamosalHornLength ~ Survival, data = lizard, var.equal = TRUE)
```

***

<br>

# Example 12.4. Salmon survival

***Welch's two-sample *t*-test for unequal variances**, comparing chinook salmon survival in the presents and absence of brook trout. This same example to demonstrate the **95% confidence interval for the ratio of two variances**, and the **F*-test of equal variances**.*

<br>

## Read and inspect data

```{r eval = TRUE}
chinook <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter12/chap12e4ChinookWithBrookTrout.csv"), stringsAsFactors = FALSE)
head(chinook)
```

<br>

## Order categories

Set the preferred order of categories in tables and graphs

```{r eval = TRUE}
chinook$troutTreatment <- factor(chinook$troutTreatment, 
	  levels = c("present", "absent"))
```

<br>

## Strip chart with error bars

Use ```stat_summary()``` to overlay mean and 95% confidence intervals. We shifted the position of the bars using ```position_nudge()``` to eliminate overlap.

fun.data=mean_cl_normal, width=0.1, conf.int=0.95

```{r eval = TRUE}
ggplot(chinook, aes(x = troutTreatment, y = proportionSurvived)) +
    geom_point(color = "firebrick", size = 3, shape = 1) +
    stat_summary(fun.data = mean_cl_normal, geom = "errorbar", 
        width = 0.1, position=position_nudge(x = 0.15)) +
    stat_summary(fun.y = mean, geom = "point", color = "firebrick",
        size = 3, position=position_nudge(x = 0.15)) +
    labs(x = "Treatment", y = "Proportion surviving") + 
    theme_classic()
```

<br>

## Statistics by group

The ```summarize()``` command from the ```dplyr``` package will generate the summary statistics in Table 12.4-3.

```{r eval = TRUE}
chinookStats <- summarize(group_by(chinook, troutTreatment), 
          Sample_mean = mean(proportionSurvived, na.rm = TRUE),
          Sample_standard_deviation = sd(proportionSurvived, na.rm = TRUE), 
          n = n())
as.data.frame(chinookStats)
```

<br>

## Welch's two-sample t-test

Use ```var.equal = FALSE``` argument of the ```t.test()``` function for a test of the difference between two means when variances are unequal.

```{r eval = TRUE}
t.test(proportionSurvived ~ troutTreatment, data = chinook, var.equal = FALSE)
```

***

<br>

# Comparing variances

*Here, we demonstrate the **95% confidence interval for the ratio of two variances**, and ***F*-test of equal variances,** using the chinook salmon experiment.*

<br>

## 95% CI for variance ratio

Warning: remember that this method is not robust to departures from assumption of normality.

```{r eval = TRUE}
var.test(proportionSurvived ~ troutTreatment, data = chinook)$conf.int
```

<br>

## *F*-test of equal variances

Warning: Remember that the *F*-test is not robust to departures from assumption of normality.

R doesn't put the larger variance on top, as in our Quick Formula Summary, but the result is the same. 

```{r eval = TRUE}
var.test(proportionSurvived ~ troutTreatment, data = chinook)
```

<br>

## Levene's test

This test of equal variances is more robust than the *F* test. The ```leveneTest()``` function is in the ```car``` package.

```{r eval = TRUE}
leveneTest(chinook$proportionSurvived, group = chinook$troutTreatment, 
	center = mean)
```
