---
title: "R code for examples in Chapter 13: <br>Handling violations of assumptions"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

*** 

We used R to analyze all examples in Chapter 13. We put the code here so that you can too.

<br>
Also see our [lab on comparing two means using R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Comparing_means_of_2_groups.html).

<br>

# New methods on this page

<ul>
<li>Data transformations:

```{r eval = FALSE}
marine$lnBiomassRatio <- log(marine$biomassRatio)
```

<li>Quantile plots:

```{r eval = FALSE}
qqnorm(marine$biomassRatio, datax = TRUE)
```

<li>Mann-Whitney *U*-test (R uses the equivalent Wilcoxon rank-sum test):

```{r eval = FALSE}
wilcox.test(timeToMating ~ feedingStatus, data = cannibalism)
```

<li>Other new methods:
<ul>
<li>Shapiro-Wilk test of normality.
<li>Sign test.
<li>Permutation test.
</ul>
</ul>

***

# Load required packages 

We'll use the ```ggplot2``` add on package to make graphs, and ```dplyr``` to calculate summary statistics by group.

```{r message = FALSE}
library(ggplot2)
library(dplyr)
```

***

<br>

# Example 13.1. Marine reserves

***The normal quantile plot**, **Shapiro-Wilk test of normality**, and the **log transformation**, investigating the ratio of biomass between marine reserves and non-reserve control areas*.

<br>

## Read and inspect data

```{r eval = TRUE}
marine <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter13/chap13e1MarineReserve.csv"), stringsAsFactors = FALSE)
head(marine)
```

<br>

## Histogram

A histogram visualizes the frequency distribution of biomass ratio (Figure 13.1-4).

```{r eval = TRUE}
hist(marine$biomassRatio, right = FALSE, col = "firebrick", 
     las = 1, xlab = "Biomass ratio", main = "")
```

or

```{r eval = TRUE}
ggplot(marine, aes(biomassRatio)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.5, 
        boundary = 0, closed = "left") + 
    labs(x = "Biomass ratio", y = "Frequency") + 
    theme_classic()
```

<br>

## Normal quantile plot

The line in the normal quantile plot of Figure 13.1-4 passes through the first and 3rd quartiles of the data.

```{r eval = TRUE}
qqnorm(marine$biomassRatio, col = "firebrick", pch = 16, las = 1,
       datax = TRUE, xlab = "Biomass ratio")
qqline(marine$biomassRatio, datax = TRUE, lwd = 2)
```

<br>

## Shapiro-Wilk test

A test of the goodness of fit of the normal distribution to data.

```{r eval = TRUE}
shapiro.test(marine$biomassRatio)
```

<br>

## Natural log-transformation

The function ```log()``` takes the natural logarithm of all the elements of a vector or variable. The following command saves the result as a new variable in the original data frame, ```marine```.

```{r eval = TRUE}
marine$lnBiomassRatio <- log(marine$biomassRatio)
```

<br>

## Histogram

A histogram of the data after log-transformation of marine biomass ratio (Figure 13.3-2).

```{r eval = TRUE}
hist(marine$lnBiomassRatio, right = FALSE, col = "firebrick", 
     las = 1, xlab = "Log biomass ratio", main = "",
     breaks = seq(-0.25, 1.5, by = 0.25))
```

or

```{r eval = TRUE}
ggplot(marine, aes(lnBiomassRatio)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.25, 
        boundary = 0, closed = "left") + 
    labs(x = "Log biomass ratio", y = "Frequency") + 
    theme_classic()
```

<br>

## 95% CI of the mean

Confidence interval for the mean of the log-transformed data.

```{r eval = TRUE}
t.test(marine$lnBiomassRatio)$conf.int
```

<br>

## Back-transform 95% CI

Transform the lower and upper limits of the confidence interval back to the original untransformed scale (```exp()``` is the inverse of the natural log function).

```{r eval = TRUE}
exp( t.test(marine$lnBiomassRatio)$conf.int )
```

***

<br>

# Example 13.4. Sexual conflict

***The sign test**, comparing the numbers of species in 25 pairs of closely related insect taxa*.

<br>

## Read and inspect data.

```{r eval = TRUE}
conflict <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter13/chap13e4SexualConflict.csv"), stringsAsFactors = FALSE)
head(conflict)
```

<br>

## Histogram

A histogram of the difference in numbers of species (Figure 13.4-1).

```{r eval = TRUE}
hist(conflict$difference, right = FALSE, col = "firebrick",
	breaks = 50, xlab = "Difference in species number", las = 1)
```

<br>

## Frequencies

Count up the frequency of differences that are below, equal to, and above zero. 

The command below creates a new variable in the data frame called ```conflictZero``` to indicate whether each difference is below, equal to, or above 0. 

```{r eval = TRUE}
conflictZero <- cut(conflict$difference, breaks = c(-Inf, 0, 1, Inf), 
    labels = c("below","equal","above"), right = FALSE)
table(conflictZero)
```

<br>

## Sign test

The sign test is just a binomial test. The output includes a confidence interval for the proportion using the Clopper-Pearson method, which isn't covered in the book.

```{r eval = TRUE}
binom.test(7, n = 25, p = 0.5)
```

***

<br>

# Example 13.5. Cricket cannibalism

***The Wilcoxon rank-sum test (equivalent to the Mann-Whitney *U*-test)** comparing times to mating (in hours) of starved and fed female sagebrush crickets. We also apply the **permutation test** to the same data*.

<br>

## Read and inspect data

```{r eval = TRUE}
cannibalism <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter13/chap13e5SagebrushCrickets.csv"), stringsAsFactors = FALSE)
head(cannibalism)
```

<br>

## Multiple histograms

The following commands produce Figure 13.5-1. We begin by ordering the categories of the feeding status variable so that "starved" comes before "fed" in the histogram.

```{r eval = TRUE}
cannibalism$feedingStatus <- factor(cannibalism$feedingStatus, 
    levels = c("starved", "fed"))

ggplot(cannibalism, aes(x = timeToMating)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 20, 
        boundary = 0, closed = "left") +
    facet_wrap( ~ feedingStatus, ncol = 1, scales = "free_y") +
    labs(x = "Time to mating (hours)", y = "Frequency") + 
    theme_classic()
```

<br>

## Wilcoxon rank-sum test

This test, available in R, is equivalent to the Mann-Whitney *U*-test.

```{r eval = TRUE}
wilcox.test(timeToMating ~ feedingStatus, data = cannibalism)
```

***

<br>

# Permutation test

***Permutation test** of the difference between mean time to mating of starved and fed crickets*.

<br>

## Observed value

Begin by calculating the **observed difference between means** (starved minus fed). The difference is -18.25734 in this data set.

```{r eval = TRUE}
cricketMeans <- summarize(group_by(cannibalism, feedingStatus), 
                  timeToMating = mean(timeToMating, na.rm = TRUE))
cricketMeans <- as.data.frame(cricketMeans)
cricketMeans

diffMeans <- cricketMeans$timeToMating[1] - cricketMeans$timeToMating[2]
diffMeans
```

<br>

## Number of permutations

Decide on the number of permutations. We use 10K here. 

```{r eval = TRUE}
nPerm <- 10000
```

<br>

## Run a loop

Create a loop to permute the data many times (the number of times determined by ```nperm```). In the following loop, ```i``` is a counter that climbs from 1 to ```nPerm``` by 1. In each iteration, the permuted difference is saved in the object ```permResult```.

```{r eval = TRUE}
permResult <- vector() # initializes
for(i in 1:nPerm){
	# step 1: permute the times to mating
	cannibalism$permSample <- sample(cannibalism$timeToMating, replace = FALSE)
	# step 2: calculate difference betweeen means
	permSampleMean <- as.data.frame(summarize(group_by(cannibalism, feedingStatus), 
                  permMean = mean(permSample, na.rm = TRUE)))
	permResult[i] <- permSampleMean$permMean[1] - permSampleMean$permMean[2]
	}
```

<br>

## Plot null distribution

Make a histogram of the permuted differences (Figure 13.8-1).

```{r eval = TRUE}
hist(permResult, right = FALSE, breaks = 100)
```

or

```{r eval = TRUE}
ggplot(data.frame(permResult), aes(permResult)) + 
    geom_histogram(fill = "white", col = "black", binwidth = 1, 
        boundary = 0, closed = "left") + 
    labs(x = "Permuted difference in treatment means (hours)", y = "Frequency") + 
    theme_classic()
```

<br>

## Approximate $P$-value

Use the null distribution to calculate an approximate *P*-value. This is the twice the proportion of the permuted means that fall below the observed difference in means, ```diffMeans``` (-18.25734 in this example). The following code calculates the *number* of permuted means falling below ```diffMeans```.

```{r eval = TRUE}
sum(as.numeric(permResult <= diffMeans))
```

These commands obtain the *fraction* of permuted means falling below ```diffMeans```.

```{r eval = TRUE}
sum(as.numeric(permResult <= diffMeans)) / nPerm
```

Finally, multiply by 2 to get the *P*-value for a two-sided test. The value won't be identical to the one in the book, or to the value you obtain, because 10,000 iterations is not large enough for extreme accuracy.

```{r eval = TRUE}
2 * ( sum(as.numeric(permResult <= diffMeans)) / nPerm )
```

