---
title: "R code for example in Chapter 20: <br>Likelihood"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

We used R to analyze all examples in Chapter 20. We put the code here so that you can too.

<br>

# New methods on this page

<ul>
<li>Binomial log-likelihood of a value for the probability of success:

```{r eval = FALSE}
dbinom(23, size = 32, p = 0.5, log = TRUE)
```

<li>Other new methods:
<ul>
<li>Log-likelihood curve.
<li>Log-likelihood ratio test.
<li>Likelihood-based 95% confidence interval.
</ul>
</ul>

***

# Load required packages 

We only need ```ggplot2``` for ordinary graphs.

```{r message = FALSE}
library(ggplot2)
```

***

<br>

# Example 20.3 Unruly passengers

***Maximum likelihood estimation** of the proportion of parasitic wasp individuals that choose the mated butterflies in a choice test. Below, we also apply the **log-likelihood ratio test** to these data*.

<br>

## Read and inspect data

```{r eval = TRUE}
wasps <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter20/chap20e3UnrulyPassengers.csv"))
head(wasps)
```

<br>

## Frequency table

Frequency table of wasp choice.

```{r eval = TRUE}
table(wasps)
```

<br>

## Likelihood and log-likelihood

The likelihood and log-likelihood of $p$ = 0.5.

```{r eval = TRUE}
dbinom(23, size = 32, p = 0.5)
dbinom(23, size = 32, p = 0.5, log = TRUE)
```

<br>

## Log-likelihoods of $p$

The log-likelihood of a range of different values of $p$ (Table 20.3-1) is obtained as follows.

```{r eval = TRUE}
proportion <- seq(0.1, 0.9, by = 0.1)
logLike <- dbinom(23, size = 32, p = proportion, log = TRUE)
data.frame(Proportion = proportion, Loglikelihood = logLike)
```

The log-likelihood calculated using a narrower range of values for $p$ (Table 20.3-2). The additional quantity ```dlogLike``` is the difference between each likelihood and the maximum.

```{r eval = TRUE}
proportion <- seq(0.4, 0.9, by = 0.01)
logLike <- dbinom(23, size = 32, p = proportion, log = TRUE)
dlogLike <- logLike - max(logLike)
```

Let's put the result into a data frame.

```{r eval = TRUE}
likeResults <- data.frame(Proportion = proportion, Loglikelihood = logLike, 
                      diffLogLike = dlogLike)
head(likeResults)
```

<br>

## Maximum likelihood estimate

```{r eval = TRUE}
pHat <- proportion[logLike == max(logLike)]
pHat
```

<br>

## Log-likelihood curve

The log-likelihood curve (Figure 20.3-1) is plotted below.

```{r eval = TRUE}
ggplot(data = likeResults, aes(x = Proportion, y = Loglikelihood)) +
  geom_line() +
  geom_hline(yintercept = max(logLike), linetype = "dashed") +
  geom_vline(xintercept = pHat, color = "firebrick", size = 1.5) +
  labs(y = "Log-likelihood", 
    x = expression(paste("Hypothesized proportion, ", italic(p)))) + 
  theme_classic()
```

<br> 

## Confidence interval

An approximate likelihood-based 95% confidence interval is obtained as follows.

```{r eval = TRUE}
lower <- max( proportion[proportion <= pHat & logLike <= max(logLike) - 1.92] )
upper <- min( proportion[proportion <= pHat & logLike <= max(logLike) - 1.92] )
c(lower = lower, upper = upper)
```

***

<br> 

# Log-likelihood ratio test

The **log-likelihood ratio test** to test the null hypothesis that the population proportion of wasps choosing mated female butterflies is 0.5. ```G``` is the test statistic and ```P``` is the *P*-value.

```{r eval = TRUE}
G <- 2 * ( dbinom(23, size = 32, p = pHat, log = TRUE)
	 - dbinom(23, size = 32, p = 0.5, log = TRUE) )
G
P <- 1 - pchisq(G, 1)
P
```


***

<br>

# Example 20.4 Conservation Scoop

***Maximum likelihood estimation** of elephant population size using mark-recapture data*.

<br>

## Read and inspect data

The data are on the 74 individuals collected in the second sample.

```{r eval = TRUE}
elephant <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter20/chap20e4ConservationScoop.csv"))
head(elephant)
```

<br>

## Frequency table

The frequency table of the 74 recaptured or unmarked individuals in the second sample.

```{r eval = TRUE}
table(elephant)
```

<br>

## Quantities needed

These are the quantities needed for the likelihood calculations.

```{r eval = TRUE}
m <-  27     # size of first sample (total marked individuals)
n2 <- 74     # size of second sample
Y <-  15     # number of recaptures in second sample
```

Choose a range of values of ```N``` to try. ```N``` is the (unknown) total number of individuals in the population. ```N``` must be at least ```n2 + m - Y``` (86, in the current example).

```{r eval = TRUE}
N <-  90:250 # The values of N to try
```

<br>

## Likelihood and log-likelihood

These are the likelihoods and log-likelihood of each ```N```.

```{r eval = TRUE}
like <- choose(m, Y) *  choose(N - m, n2 - Y) /  choose(N, n2)
logLike <- lchoose(m, Y) + lchoose(N - m, n2 - Y) - lchoose(N, n2)
```

Likelihood and log-likelihood can also be obtained using R's built-in function for the hypergeometric distribution.

```{r eval = TRUE}
like <- dhyper(Y, m, N - m, n2)
logLike <- dhyper(Y, m, N - m, n2, log = TRUE)
```

<br>

## Maximum likelihood estimate

The maximum likelihood estimate of elephant population size.

```{r eval = TRUE}
Nhat <- N[logLike == max(logLike)]
Nhat
```

<br>

## Log-likelihood curve

The log-likelihood curve (Figure 20.4-1) is produced as follows.

```{r eval = TRUE}
likeResults <- data.frame(N, like, logLike)
ggplot(data = likeResults, aes(x = N, y = logLike)) +
  geom_line() +
  labs(y = "Log-likelihood", 
    x = expression(paste("Population size estimate, ", italic(N)))) + 
  theme_classic()
```

<br>

## Confidence interval

The likelihood-based 95% confidence interval for elephant population size is approximately as follows.

```{r eval = TRUE}
lower <- max(N[N <= Nhat & logLike <= max(logLike) - 1.92])
upper <- min(N[N <= Nhat & logLike <= max(logLike) - 1.92])
c(lower = lower, upper = upper)
```
