---
title: "R code for example in Chapter 10: <br>The normal distribution"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

We used R to analyze all examples in Chapter 10. We put the code here so that you can too.

<br>
Also see our [lab on the normal distribution using R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Normal_and_sample_means.html).

<br>

# New methods on this page

<ul>
<li>Probabilities under the normal curve:

```{r eval = FALSE}
pnorm(157.5, mean = 177.6, sd = 9.7)
```

<li>Other new methods:
<ul>Normal approximation to a binomial distribution.
</ul>
</ul>

***

<br>

# Load required packages 

We'll use the ```ggplot2``` add on package to make graphs.

```{r message=FALSE}
library(ggplot2)
```

***

<br>

<!--
NASA excludes anyone under 62 inches in height and anyone over 75 inches from being
an astronaut pilot (NASA 2004). In metric units, these values for the lower and upper
height restrictions are 157.5 cm and 190.5 cm, respectively. What fraction of the young
adult American population is excluded from being an astronaut pilot by these height constraints?
The distribution of adult heights within a gender and age group is reasonably
well approximated by a normal distribution, with the mean and standard deviation for
20- to 29-year-old males in America given by 177.6 cm and 9.7 cm, respectively (McDowell
et al. 2008). For 20- to 29-year-old American females, the mean height is 163.2 cm with a
standard deviation of 10.1 cm.
-->

# Example 10.4. Small step for man?

***Calculate probabilities under the normal curve***. 

<br>

## Adult men

The command ```pnorm(Y)``` gives the probability of obtaining a value **less than** ```Y``` under the normal distribution. Set the arguments ```mean =``` and ```sd =``` to the mean and standard deviate of the desired normal distribution.

Here are the calculations for adult men. The numbers here will be slightly different from those in the book because of roundoff errors.

Pr[Height < 157.5]

```{r eval = TRUE}
pTooShort <- pnorm(157.5, mean = 177.6, sd = 9.7)
pTooShort
```

Pr[Height > 190.5]

```{r eval = TRUE}
pTooTall <- 1 - pnorm(190.5, mean = 177.6, sd = 9.7)
pTooTall
```

Pr[Height < 157.5 or Height > 190.5]

```{r eval = TRUE}
pTooShort + pTooTall
```


<br>

## Adult women

<!-- 
Women’s heights
are approximated by a normal distribution with mean 163.2 cm and standard
deviation 10.1 cm. These values correspond to height restrictions at 0.56 standard
deviations below the mean and 2.7 standard deviations above the mean. (Check these
numbers for yourself.) Using Statistical Table B, the fraction 0.28774 of the women
are too short to meet NASA’s guidelines and the fraction 0.00347 are too tall. In
total, the height restrictions exclude 29.1% of young American women from being
astronaut pilots.
-->

Pr[Height < 157.5]

```{r eval = TRUE}
pTooShort <- pnorm(157.5, mean = 163.2, sd = 10.1)
pTooShort
```

Pr[Height > 190.5]

```{r eval = TRUE}
pTooTall <- 1 - pnorm(190.5, mean = 163.2, sd = 10.1)
pTooTall
```

Pr[Height < 157.5 or Height > 190.5]

```{r eval = TRUE}
pTooShort + pTooTall
```

***

<br>

# Figure 10.6. Spanish flu

***Demonstration of the central limit theorem**, using the distribution of sample mean age at death in samples from a highly non-normal distribution*.

Read and inspect the data.

```{r eval = TRUE}
flu <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter10/chap10e6AgesAtDeathSpanishFlu1918.csv"), stringsAsFactors = FALSE)
head(flu)
```

<br>

## Histogram 

Visualizing the frequency distribution of ages at death in Switzerland in 1918 during the Spanish flu epidemic.

```{r eval = TRUE}
hist(flu$age, right = FALSE, col = "firebrick", breaks = seq(0,102,2), 
     xlab = "Age at death (yrs)", las = 1, main = "")
```

or

```{r eval = TRUE}
ggplot(flu, aes(x = age)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 2, 
        boundary = 0, closed = "left") + 
    labs(x = "Age at death (yrs)", y = "Frequency") + 
    theme_classic()
```

<br>

## Demonstrate central limit theorem

We treat the age at death measurements from Switzerland in 1918 as a population. Then we take a large number of random samples, each of size $n$, from the population of age at death measurements and plot the sample means. 

Note: your results won't be the identical to the corresponding panel in Figure 10.6-2, because 10,000 random samples is not large enough for extreme accuracy. 

The code below is for $n = 4$. Change $n$ to another number and rerun to see the effects of sample size on the shape of the distribution of sample means.

```{r eval = TRUE}
n <- 4
results <- vector()
for(i in 1:10000){
	AgeSample <- sample(flu$age, size = n, replace = FALSE)
	results[i] <- mean(AgeSample)
	}
```


<br>

## Histogram of sample means

```{r eval = TRUE}
hist(results, right = FALSE, breaks = 50, col = "firebrick", las = 1, 
	xlab = "Mean age at death (yrs)", ylab = "Frequency", main = "")
```

or

```{r eval = TRUE}
ggplot(data.frame(results), aes(x = results)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 2, 
        boundary = 0, closed = "left") + 
    labs(x = "Mean age at death (yrs)", y = "Frequency") + 
    theme_classic()
```

***

<br>

# Example 10.7. Dead bug

***Normal approximation to the binomial distribution** applied to the cricket preference of the brown recluse spider*.

<br>

## Normal approximation

The below differs slighty from the number in the book because of roundoff.

```{r eval = TRUE}
spiderProb <- 1 - pnorm( (30 + 1/2 - 41 * 0.5) / sqrt(41 * 0.50 * 0.5))
Pvalue <- 2 * spiderProb
Pvalue
```

<br>

## Compare with exact result

Compare with the exact result obtained when using the binomial distribution, ```dbinom()```, which we encountered in the Chapter 7 R page.

```{r eval = TRUE}
2 * sum( dbinom(31:41, size = 41, prob = 0.5) )
```

You can also use ```pbinom()```.

```{r eval = TRUE}
2 * (1 - pbinom(30, size = 41, prob = 0.5))
```

