---
title: "R code for example in Chapter 19: <br>Computer-intensive methods"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

<br>

We used R to analyze all examples in Chapter 19. We put the code here so that you can too.

# New methods on this page

<ul>
<li>Calculate a *P*-value for a χ<sup>2</sup> goodness-of-fit test using simulation:

```{r eval = FALSE}
chisq.test(observedFreq, simulate.p.value = TRUE, B = 10000)
```

<li>Calculate a bootstrap standard error:

```{r eval = FALSE}
boot(chimp$asymmetryScore, boot.median, R = 10000)
```

<li>Other new methods:
<ul>
<li>Bootstrap 95% confidence interval.
</ul>
</ul>
 
***

# Load required packages 

We need ```ggplot2``` for ordinary graphs. The ```boot``` package is for bootstrap estimation.

```{r message = FALSE}
library(ggplot2)
library(boot)
```

***

<br>

# Example 19.1 Two-digit "psychic"

***Hypothesis testing using simulation** to determine whether all two-digit numbers occur with equal probability when people choose two-digit numbers haphazardly*.

## Read and examine data

Each row is the two-digit number chosen by a different individual.

```{r eval = TRUE}
haphazard <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter19/chap19e1TwoDigitNumbers.csv"), stringsAsFactors = FALSE)
head(haphazard)
```

<br>

## Histogram

Plot the frequency distribution of two-digit numbers chosen by volunteers (Figure 19.1-1).

```{r eval = TRUE}
ggplot(data = haphazard, aes(x = numberSelected)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 1, 
        closed = "left", boundary = 0) + 
    scale_x_continuous(breaks = seq(from = 10, to = 100, by = 10)) +
    labs(x = "Number thought of by volunteer", y = "Frequency") + 
    theme_classic()
```

<br>

## Frequency table

Table the two-digit numbers chosen by volunteers. Notice that many numbers between 0 and 99 are not represented in the data. 

```{r eval = TRUE}
table(haphazard$numberSelected)
```

To make sure that numbers not chosen are nevertheless included in the calculations, we must convert the variable to a factor and specify all possible categories.

```{r eval = TRUE}
factorData <- factor(haphazard$numberSelected, levels = 10:99)
observedFreq <- table(factorData)
observedFreq
```

<br>

## Test statistic 

Here, we are using the $\chi^2$ goodness-of-fit statistic. We are lazy and use the ```chisq.test()``` function to do the calculations. R will give you a warning because of the low expected frequencies. This is exactly why we need to use simulation instead of the $\chi^2$ goodness-of-fit-test.

```{r eval = TRUE}
chiData <- chisq.test(observedFreq)$statistic
chiData
```

<br>

## Simulate a single random sample

We'll need to know the sample size (number of rows, in this case) to run the simulation.

```{r eval = TRUE}
n <- nrow(haphazard)
n
```

Simulate $n$ numbers from 10 to 99 (with replacement), where $n$ is the number of individuals in the sample.

```{r eval = TRUE}
randomTwoDigit <- sample(10:99, size = n, replace = TRUE)
```

<br>

## Statistic for simulated sample

Remember to convert data to a factor to make sure that all categories are included in the calculations. Under the null hypothesis, the expected frequency is the same for all numbers.

```{r eval = TRUE}
randomTwoDigit <- factor(randomTwoDigit, levels = 10:99)
observedFreq <- table(randomTwoDigit)
expectedFreq <- n/length(10:99)
chiSim <- sum( (observedFreq - expectedFreq)^2 / expectedFreq )
chiSim
```

<br>

## Repeat many times

The following is a loop repeated ```nSim``` times. In each iteration ```i```, a random sample of numbers is simulated, a frequency table is calculated, and the $\chi^2$ statistic is computed. The statistic is saved in the ```i```<sup>th</sup> element of the vector ```results```.

```{r eval = TRUE}
nSim <- 10000
results <- vector()
for(i in 1:nSim){
	randomTwoDigit <- sample(10:99, size = n, replace = TRUE)
	randomTwoDigit <- factor(randomTwoDigit, levels = 10:99)
	simFreq <- table(randomTwoDigit)
	expectedFreq <- n/length(10:99)
	results[i] <- sum( (simFreq - expectedFreq)^2 / expectedFreq )
	}
```

<br>

## Null distribution of $\chi^2$

Plot the frequency distribution of $\chi^2$ values from the simulation (Figure 19.1-2). This is the simulated null distribution for the test statistic.

```{r eval = TRUE}
nullData <- data.frame(results)
ggplot(data = nullData, aes(x = results)) + 
    geom_histogram(fill = "white", col = "black", binwidth = 2, 
        closed = "left", boundary = 0) + 
    scale_x_continuous(breaks = seq(from = 40, to = 160, by = 20)) +
    labs(x = expression(paste("Simulated ", chi^2)), y = "Frequency") + 
    theme_classic()
```

<br>

## *P*-value

The $P$-value is the fraction of simulated $\chi^2$ values equalling or exceeding the observed value of the test statistic in the data.

```{r eval = TRUE}
P <- sum(results >= chiData)
P
```

<br>

## Built-in method

R also has a built-in method to simulate the null distribution and calculate the $P$-value. Using it simply involves providing a couple of arguments to the ```chisq.test()``` function. ```B``` is the number of iterations desired. If you don't provide the probabilities of each outcome under the null hypothesis as an argument, R assumes that all the categories are equiprobable under the null hypothesis.

```{r eval = TRUE}
factorData <- factor(haphazard$numberSelected, levels = 10:99)
observedFreq <- table(factorData)
chisq.test(observedFreq, simulate.p.value = TRUE, B = 10000)
```

***

<br>

# Example 19.2 Chimp language centers

***Bootstrap estimation of the median** asymmetry score for Brodmann’s area 44 in chimpanzees*.

<br>

## Read and examine data

```{r eval = TRUE}
chimp <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter19/chap19e2ChimpBrains.csv"), stringsAsFactors = FALSE)
head(chimp)
```

<br>

## Histogram

Histogram of asymmetry scores of the 20 chimps (Figure 19.2-1).

```{r eval = TRUE}
ggplot(data = chimp, aes(x = asymmetryScore)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.25, 
        closed = "left", boundary = 0) + 
    labs(x = "Brodmann‘s area 44 asymmetry score", y = "Frequency") + 
    theme_classic()
```

<br>

## Test statistic

Calculate the median asymmetry score from the data. The median is our test statistic.

```{r eval = TRUE}
median(chimp$asymmetryScore)
```

<br>

## Single bootstrap replicate

Obtain a single bootstrap replicate. This involves sampling (with replacement) from the actual data. Calculate the median of this bootstrap replicate.

```{r eval = TRUE}
bootSample <- sample(chimp$asymmetryScore, size = nrow(chimp), replace = TRUE)
median(bootSample)
```

<br>

## Repeat many times

Repeat this process many times. The following is a loop repeated ```B``` times. In each iteration, the bootstrap replicate estimate (median) is recalculated and saved in the results vector ```bootMedian```.

```{r eval = TRUE}
B <- 10000
bootMedian <- vector()
for(i in 1:B){
	bootSample <- sample(chimp$asymmetryScore, size = nrow(chimp), replace = TRUE)
	bootMedian[i] <- median(bootSample)
	}
```

<br>

## Bootstrap replicate estimates

Draw a histogram of the bootstrap replicate estimates for median asymmetry (Figure 19.2-2). 

Note: your results won't be the identical to the one in Figure 19.2-2, because 10,000 random samples is not large enough to obtain the sampling distribution with extreme accuracy. Increase ```B``` in the above loop for greater accuracy.

```{r eval = TRUE}
bootResults <- data.frame(bootMedian)
ggplot(data = bootResults, aes(x = bootMedian)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.01, 
        closed = "left", boundary = 0) + 
    labs(x = "Bootstrap replicate estimates of the median asymmetry", y = "Frequency") + 
    theme_classic()
```

<br>

## Mean of bootstrap replicate estimates

Calculate the mean of the bootstrap replicate estimates.

```{r eval = TRUE}
mean(bootMedian)
```

<br>

## Bootstrap standard error

The bootstrap standard error is the standard deviation of the bootstrap replicate estimates.

```{r eval = TRUE}
sd(bootMedian)
```

<br> 

## Bootstrap confidence interval

Use the percentiles of the bootstrap replicate estimates to obtain a bootstrap 95% confidence interval for the population median.

```{r eval = TRUE}
quantile(bootMedian, probs = c(0.025, 0.975))
```

<br>

## Use ```boot``` package

Use the ```boot``` package in R for bootstrap estimation.

To begin, we need to define a new R function to tell the ```boot``` package what statistic we want to estimate, which is the median in this example. Let's call the new function ```boot.median()```. The syntax required by the package is not too complex, but our function must include a counter as an argument (as usual, we call it ```i```).

```{r eval = TRUE}
boot.median <- function(x, i){ median(x[i]) }
```

Use the ```boot()``` function of the ```boot``` package to obtain many bootstrap replicate estimates and calculate the bootstrap standard error.

```{r eval = TRUE}
bootChimp <- boot(chimp$asymmetryScore, boot.median, R = 10000)
bootChimp
```

<br>

## Histogram of bootstrap replicate estimates

```{r eval = TRUE}
medianBootChimp <- data.frame(medianChimp = bootChimp$t)
ggplot(data = medianBootChimp, aes(x = medianChimp)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.01, 
        closed = "left", boundary = 0) + 
    labs(x = "Bootstrap replicate estimates of the median asymmetry", y = "Frequency") + 
    theme_classic()
```

<br>

## Bootstrap confidence interval

Get the bootstrap 95% confidence interval for the population median. We show two methods. The percentile method is already familiar, but ```boot``` also provides an improved method, called "bias corrected and accelerated".

```{r eval = TRUE}
boot.ci(bootChimp, type = "perc")
boot.ci(bootChimp, type = "bca")
```

***

<br>

# Figure 19.2-3 Sexual cannibalism revisited

***Bootstrap estimation of the difference between medians of two groups**, using times to mating (in hours) of female sagebrush crickets that were either starved or fed (data are from Example 13.5). We show the method **using R's ```boot``` package***.

<br>

## Read and inspect data

```{r eval = TRUE}
cannibalism <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter13/chap13e5SagebrushCrickets.csv"), stringsAsFactors = FALSE)
head(cannibalism)
```

Calculate the **observed difference** between the sample medians of the two groups.

```{r eval = TRUE}
twoMedians <- tapply(cannibalism$timeToMating, cannibalism$feedingStatus, median)
diffMedian <- twoMedians[2] -  twoMedians[1]
diffMedian
```

<br>

## Function to calculate test statistic

Define a new function to calculate the difference between the medians of the bootstrap replicate estimates. Let's call it ```boot.diffMedian```.

```{r eval = TRUE}
boot.diffMedian <- function(x, i){
	twoMedians <- tapply(x$timeToMating[i], x$feedingStatus[i], median)
	diffMedian <- twoMedians[2] -  twoMedians[1]
	}
```

<br>

## Run ```boot()```

Obtain the many bootstrap replicate estimates and get the bootstrap standard error.

```{r eval = TRUE}
bootResults <- boot(cannibalism, boot.diffMedian, R = 10000)
bootResults
```

<br>

## Histogram of bootstrap replicate estimates

Histogram of the difference between two medians (Figure 19.2-3).

```{r eval = TRUE}
diffBootMedians <- data.frame(diffMedians = bootResults$t)
ggplot(data = diffBootMedians, aes(x = diffMedians)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 5, 
        closed = "left", boundary = 0) + 
    labs(x = "Bootstrap replicate difference in median", y = "Frequency") + 
    theme_classic()
```

<br>

## Bootstrap confidence interval

Obtain the bootstrap 95% confidence interval for the difference between the population medians, using the percentile and the "bca" method.

```{r eval = TRUE}
boot.ci(bootResults, type = "perc")
boot.ci(bootResults, type = "bca")
```
