---
title: "R code for examples in Chapter 3: <br>Describing data"
author:

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
```

***

We used R to analyze all examples in Chapter 3. We put the code here so that you can too.

<br>
Also see our [lab on describing data using R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Describing_data.html).

<br>

# New methods on this page

<ul>
<li> Sample mean, when there are no missing ```(NA)``` entries:

```{r eval = FALSE}
mean(snakeData$undulationRate)
```

<li>Standard deviation, when there are no missing ```(NA)``` entries:

```{r eval = FALSE}
sd(snakeData$undulationRate)
```

<li>Calculate a statistic by group, when there are no missing ```(NA)``` entries:

```{r eval = FALSE}
summarize(group_by(sticklebackData, genotype), Mean = mean(plates))
```

<li>Extract a subset of data from a data frame:

```{r eval = FALSE}
subset(spiderData, treatment == "before")
```


<li>Other new methods:

<ul>
<li>Variance and coefficient of variation.
<li>Median and interquartile range.
<li>Round to a preferred number of decimals.
<li>Cumulative frequency distribution.
<li>Table of descriptive statistics by group.
<li>Mean and standard deviation from a frequency table.
<li>Order categories of a categorical variable in tables and graphs.
</ul>
</ul>

***

<br>

# Load required packages

We'll use the ```ggplot2``` package to make plots and the ```dplyr``` package to tabulate summary statistics.

```{r eval = TRUE}
library(ggplot2)
library(dplyr)
```

***

<br>

# Example 3.1-1. Gliding snakes

***Sample mean**, **standard deviation**, **variance** and **coefficient of variation** of undulation rate of 8 gliding paradise tree snakes, in Hz*.

<br> 

## Read and inspect data

```{r eval = TRUE}
snakeData <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter03/chap03e1GlidingSnakes.csv"), stringsAsFactors = FALSE)

head(snakeData)
```

<br>

## Histogram 

This is the histogram of Figure 3.1-1.

```{r eval = TRUE}
ggplot(data = snakeData, aes(x = undulationRateHz)) + 
    geom_histogram(fill = "firebrick", col = "black", 
          breaks = seq(0.8, 2.2, by = 0.2), boundary = 0, closed = "left") + 
    labs(x = "Undulation rate (Hz)", y = "Frequency") + 
    theme_classic()
```

<br> 

## Mean, standard deviation, variance

Calculate the **mean, standard deviation** and **variance** of the undulation rates.

```{r eval = TRUE}
mean(snakeData$undulationRate)
sd(snakeData$undulationRate)
var(snakeData$undulationRate)
```

<br>

## Coefficient of variation.

```{r eval = TRUE}
100 * sd(snakeData$undulationRate)/mean(snakeData$undulationRate)
```

***

<br>

# Table 3.1-2. Numbers of convictions

<i>**Mean and standard deviation from a frequency table**. The data are from Chapter 2</i>.

<br>

## Read and inspect the data. 

The data are stored as a frequency table of the number of convictions.

```{r eval = TRUE}
convictionsFreq <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter03/chap03t1_2ConvictionsFreq.csv"), stringsAsFactors = FALSE)
head(convictionsFreq)
```

<br>

## ```rep()``` command

First, use the ```rep()``` command to repeat each value in the table, the number of times according to frequency. Here, we store the result in ```convictions```.

```{r eval = TRUE}
convictions <- rep(convictionsFreq$convictions, convictionsFreq$frequency)
```

<br> 

## Mean and standard deviation

```{r eval = TRUE}
mean(convictions)
sd(convictions)
```

These operations won't work if there are missing values in the data set. If missing values are present it is necessary to add an argument to the command as follows. 

```{r eval = TRUE}
mean(convictions, na.rm = TRUE)
sd(convictions, na.rm = TRUE)
```

***

<br>

# Example 3.2. Spider running speed

***Median**, **interquartile range**, and **box plot** of running speed (cm/s) of male <i>Tidarren</i> spiders.
*

<br>

## Read and inspect data

The data are in "long" format. One variable indicates running speed, and a second variable gives treatment (before vs after amputation). Therefore, every individual spider is on two rows, once for its before-amputation measurement and one for its after-amputation measurement.

```{r eval = TRUE}
spiderData <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter03/chap03e2SpiderAmputation.csv"), stringsAsFactors = FALSE)
head(spiderData)
```

<br>

## Box plot

Begin by ordering the treatment levels so that the "before" amputation measurements come before the "after" measurements in the plot.

```{r eval = TRUE}
spiderData$treatment <- factor(spiderData$treatment, levels = c("before", "after"))

ggplot(spiderData, aes(x = treatment, y = speed)) + 
    geom_boxplot(fill = "goldenrod1") +
    theme_classic()
```

<br>

## Sample median

First, extract the before-amputation data using ```subset()```. Note the double equal sign "==" needed in the logical statement, which indicates "is equal to" in R. Save the result in a new data frame named ```speedBefore```.

```{r eval = TRUE}
speedBefore <- subset(spiderData, treatment == "before") 
speedBefore
```

## Sample median

```{r eval = TRUE}
median(speedBefore$speed)
```

<br>

## First and third quartiles

Calculate the first and third quartiles of before-amputation running speed (0.25 and 0.75 quantiles). Type = 5 reproduces the method we use in the book to calculate quartiles.

```{r eval = TRUE}
quantile(speedBefore$speed, probs = c(0.25, 0.75), type = 5)
```

<br>

## Interquartile range

Calculate the interquartile range of before-amputation running speed. The argument ```type = 5``` reproduces the method we use in the book to calculate quartiles.

```{r eval = TRUE}
IQR(speedBefore$speed, type = 5)
```

***

<br>

# Example 3.3. Stickleback lateral plates

*Draw **multiple histograms** and **produce a table of descriptive statistics by group** for plate numbers of three stickleback genotypes. We also include a **table of frequencies and proportions** of stickleback genotypes.*

<br>

## Download and inspect the data.

```{r eval = TRUE}
sticklebackData <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter03/chap03e3SticklebackPlates.csv"), stringsAsFactors = FALSE)
head(sticklebackData)
```

<br>

## Order genotypes

Begin by setting the preferred order of the three genotypes in the figure.

```{r eval = TRUE}
sticklebackData$genotype <- factor(sticklebackData$genotype, 
	  levels = c("MM","Mm","mm"))
```

<br>

## Multiple histograms

One histogram per genotype, stacked vertically, all using the same x-axis scale.

```{r eval = TRUE}
ggplot(sticklebackData, aes(x = plates)) + 
    geom_histogram(fill = "firebrick", col = "black", 
        boundary = 0, closed = "left", binwidth = 2) +
    labs(x = "Number of lateral plates", y = "Frequency") + 
    facet_wrap( ~ genotype, ncol = 1, scales = "free_y") +
    theme_classic()
```

<br>

## Statistics by group

We'll use the ```summarize()``` command from the ```dplyr``` package. The commands below assume that the variables contain no missing ```(NA)``` elements. The ```n()``` function calculates the number of cases. The function ```round(x, digits = 1)``` rounds the contents of x to the desired number of decimal places (here, 1). 

```{r eval = TRUE}
sticklebackTable <- summarize(group_by(sticklebackData, genotype), 
    n = n(),
    Mean = round( mean(plates), digits = 1 ),
    Median = median(plates),
    Std.dev = round( sd(plates), digits = 1),
    Interq.range = IQR(plates, type = 5))

data.frame(sticklebackTable)
```

<br>

# Figure 3.4-1. Spider running speed

**Draw a cumulative frequency distribution** of running speed before amputation.

<br>

## Cumulative frequency distribution

```{r eval = TRUE}
ggplot(speedBefore, aes(x = speed)) + 
	stat_ecdf(color = "firebrick", size = 1.5) +
	labs(x = "Running speed before amputation (cm/s)", 
		y = "Cumulative relative frequency") + 
	theme_classic()
```

<br>

# Table 3.5-1. Genotype frequencies

**Make a table of frequencies and proportions** of the stickleback genotypes. Here we used ```summarize()``` from the ```dplyr``` package

<br>

## Make table

```{r eval = TRUE}
sticklebackTable <- summarize(group_by(sticklebackData, genotype), 
    n = n())
sticklebackTable$Proportion <- sticklebackTable$n/sum(sticklebackTable$n)
data.frame(sticklebackTable)
```

The table would look even better if you rounded the proportions (give this a try).

