---
title: "R code for example in Chapter 17: <br>Regression"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

We used R to analyze all examples in Chapter 16. We put the code here so that you can too.

Also see our [R lab on linear regression
](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Correlation_Regression.html).

<br>

# New methods on this page

<ul>
<li>Linear regression:
```{r eval = FALSE}
lionRegression <- lm(ageInYears ~ proportionBlack, data = lion)
summary(lionRegression)
```

<li>95% confidence interval for the slope:
```{r eval = FALSE}
confint(lionRegression)
```

<li>Predict mean y on a regression line corresponding to a given x-value:
```{r eval = FALSE}
predict(lionRegression, data.frame(proportionBlack = 0.50), se.fit = TRUE)
```

<li>ANOVA test of zero regression slope:
```{r eval = FALSE}
anova(prairieRegression)
```

<li>Other new methods:
<ul>
<li>95% confidence bands.
<li>95% prediction intervals.
<li>Scatter plot smoothing.
<li>Nonlinear regression.
<li>Logistic regression.
</ul>
</ul>
 
***

# Load required packages 

If you want to try the ```geom_spline()``` function for scatter plot smoothing using splines, you need to install the ```ggformula``` package on your computer. Load the ```MASS``` package for logistic regression.

```{r eval = FALSE}
install.packages("binom", dependencies = TRUE) # optional
```

Load the ```ggplot2``` package for graphs.

```{r message = FALSE}
library(ggplot2)
library(ggformula) # optional
library(MASS)
```

***

<br>
 
# Example 17.1. Lion noses

***Estimate a linear regression** and **calculate predicted values and their uncertainty** for the relationship between lion age and proportion of black on the lion nose*.

<br>

## Read and inspect data

```{r eval = TRUE}
lion <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17e1LionNoses.csv"), stringsAsFactors = FALSE)
head(lion)
```

<br>

## Scatter plot

This command produces the plot in Figure 17.1-1.

```{r eval = TRUE}
ggplot(lion, aes(proportionBlack, ageInYears)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Proportion black", y = "Age (years)") + 
    theme_classic()
```

<br>

## Fit regression line

Fit the linear regression to the data using least squares. Use ```lm()```, which was also used for ANOVA. Save the results into a model object, and then use other commands to extract the quantities wanted.

```{r eval = TRUE}
lionRegression <- lm(ageInYears ~ proportionBlack, data = lion)
```

Slope and intercept of the best fit line, along with their standard errors, are extracted with the ```summary()``` function. Extract just the coefficients with ```coef(lionRegression)```.

```{r eval = TRUE}
summary(lionRegression)
```

<br>

## 95& CI for intercept and slope

```{r eval = TRUE}
confint(lionRegression)
```

<br>

## Add line to scatter plot

Overlay the least squares regression line onto the plot (Figure 17.1-4) using ```geom_smooth()```.

```{r eval = TRUE}
ggplot(lion, aes(proportionBlack, ageInYears)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_smooth(method = "lm", se = FALSE, col = "black") +
    labs(x = "Proportion black", y = "Age (years)") + 
    theme_classic()
```

<br>

## Use line to predict

Use the regression line for prediction. For example, here is how to predict mean lion age corresponding to a value of 0.50 of proportion black in the nose.

```{r eval = TRUE}
yhat <- predict(lionRegression, data.frame(proportionBlack = 0.50), se.fit = TRUE)
data.frame(yhat)
```

In the output, predicted age is indicated by ```fit``` and the standard error of the predicted mean age is ```se.fit```

<br>

## 95% confidence bands

The 95% confidence bands measure uncertainty of the predicted mean values on the regression line corresponding to each x-value (Figure 17.2-1, left). The confidence limits are the upper and lower edge of the shaded area in the following plot.

```{r eval = TRUE}
ggplot(lion, aes(proportionBlack, ageInYears)) + 
    geom_smooth(method = "lm", se = TRUE, col = "black") +
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Proportion black", y = "Age (years)") + 
    theme_classic()
```

<br>

## 95% Prediction interval

The 95% prediction intervals show the 95% confidence intervals for the predicted value of a new (future) individual, given its value for the x-variable (Figure 17.2-1, right). The prediction interval is wider than the confidence bands because predicting the age of a new individual lion age from the proportion black in its nose has higher uncertainty than predicting the mean age of all lions having that proportion of black on their noses.

```{r eval = TRUE}
lion2 = data.frame(lion, predict(lionRegression, interval = "prediction"))
ggplot(lion2, aes(proportionBlack, ageInYears)) + 
    geom_ribbon(aes(ymin = lwr, ymax = upr, fill='prediction'), 
        fill = "black", alpha = 0.2) +
    geom_smooth(method = "lm", se = FALSE, col = "black") +
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Proportion black", y = "Age (years)") + 
    theme_classic()
```

***

<br>
 
# Example 17.3. Prairie home campion

***Test the null hypothesis of zero regression slope**. The data are from an experiment investigating the effect of plant species diversity on the stability of plant biomass production*.

<br>
 
## Read and inspect data

```{r eval = TRUE}
prairie <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17e3PlantDiversityAndStability.csv"), stringsAsFactors = FALSE)
head(prairie)
```

<br>
 
## Log-transform

Log-transform stability and include the new variable in the data frame. Inspect the data frame once again to make sure the function worked as intended.

```{r eval = TRUE}
prairie$logStability <- log(prairie$biomassStability)
head(prairie)
```

<br>
 
## Scatter plot with line

These commands produce the plot Figure 17.3-1.

```{r eval = TRUE}
ggplot(prairie, aes(nSpecies, logStability)) + 
    geom_point(size = 3, col = "firebrick", shape = 1) + 
    geom_smooth(method = "lm", se = FALSE, col = "black") +
    xlim(0,16) +
    labs(x = "Species number treatment", 
        y = "Log-transformed ecosystem stability") + 
    theme_classic()
```

<br>

## Least squares regression

Fit the least squares linear regression to the data using ```lm()```. Save the results into a model object, and then use ```summary()``` to extract the intercept and slope of the fitted regression line.

```{r eval = TRUE}
prairieRegression <- lm(logStability ~ nSpecies, data = prairie)
summary(prairieRegression)
```

<br>
 
## $t$-test of regression slope

This is provided as part of the output of the ```summary()``` command immediately above. 

<br>
 
## 95% CI of regression slope

```{r eval = TRUE}
confint(prairieRegression)
```

<br>
 
## ANOVA test

This method produces the ANOVA table (Table 17.3-2), which includes a test of the null hypothesis that the regression slope is zero.

```{r eval = TRUE}
anova(prairieRegression)
```

<br>
 
## $R^2$

The $R^2$ associated with the regression is part of the output of the ```summary``` command.

```{r eval = TRUE}
summary(prairieRegression)$r.squared
```

<!--
***

<br>

# Figure 17.5-2. Tail feather outlier

***Examine the effect on an outlier on linear regression**. The data are percentage of white in the tail feathers of the dark-eyed junco at sites at different latitudes*.

Read and inspect data.

```{r eval = TRUE}
junco <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f5_2JuncoOutlier.csv"), stringsAsFactors = FALSE)
head(junco)
```

**Scatter plot** of all the data.

```{r eval = TRUE}
plot(percentWhiteTail ~ latitude, data = junco)
```

Commands for a more elaborate scatter plot using more options (Figure 17.5-2) are here

```{r eval = TRUE}
plot(percentWhiteTail ~ latitude, data = junco, pch = 16, col="firebrick", 
 	las = 1, cex = 1.5, bty = "l", xlab = "Latitude (degrees N)", 
 	ylab = "Percent white in tail")
```


Add regression lines to the scatter plot, calculated both with and without the outlier. You may need to redraw the scatter plot using the above commands before executing the following functions.

```{r eval = TRUE}
juncoRegression <- lm(percentWhiteTail ~ latitude, data = junco)
plot(percentWhiteTail ~ latitude, data = junco, pch = 16, col="firebrick", 
 	las = 1, cex = 1.5, bty = "l", xlab = "Latitude (degrees N)", 
 	ylab = "Percent white in tail")
abline(juncoRegression, col = "red", lwd = 2)

```


```{r eval = TRUE}
juncoRegressionNoOut <- lm(percentWhiteTail ~ latitude, 
	data = junco, subset = percentWhiteTail < 38)
plot(percentWhiteTail ~ latitude, data = junco, pch = 16, col="firebrick", 
 	las = 1, cex = 1.5, bty = "l", xlab = "Latitude (degrees N)", 
 	ylab = "Percent white in tail")
##abline(juncoRegressionNoOut, lwd = 2) ###Mike: this line is not rendering well. I think its mostly because the plot is done on all data while the regression is only on a subset. 
```
-->

***

<br>

# Figure 17.5-3. Detecting nonlinearity

***Detect a nonlinear relationship graphically**, using data on surface area of desert pools and the number of fish species present*.

<br>

## Read and inspect data

```{r eval = TRUE}
desertFish <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f5_3DesertPoolFish.csv"), stringsAsFactors = FALSE)
head(desertFish)
```

<br>

## Scatter plot with line

The plot reveals the poor fit of a straight line (Figure 17.5-3 left).

```{r eval = TRUE}
ggplot(desertFish, aes(poolArea, nFishSpecies)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_smooth(method = "lm", se = FALSE, col = "black") +
    labs(x = expression(Area~(m^2)), y = "Number of species") + 
    theme_classic()
```

<br>

## Scatter plot smoothing

This code for Figure 17.5-3 (right) uses ```geom_spline()```, which requires the ```ggformula``` package, which extends ```ggplot()```. The ```df```option controls the degree of smoothing: a larger number results in a more wiggly curve. Otherwise try using ```geom_smooth(method = "loess") instead.

```{r eval = TRUE}
ggplot(desertFish, aes(poolArea, nFishSpecies)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_spline(col = "black", df = 2) +
    labs(x = expression(Area~(m^2)), y = "Number of species") + 
    theme_classic()
```

<br>

## Log-transformed data (Figure 17.6-2)

We log-transformed both variables and saved the results (```ln_nFishSpecies```, ```ln_poolArea```) in the data frame. Then we made a new scatter plot with the log-transformed variables and added the line to the plot.

```{r eval = TRUE}
desertFish$ln_nFishSpecies <- log(desertFish$nFishSpecies)
desertFish$ln_poolArea <- log(desertFish$poolArea)
ggplot(desertFish, aes(ln_poolArea, ln_nFishSpecies)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_smooth(method = "lm", se = FALSE, col = "black") +
    labs(x = "Ln(area)", y = "Ln(number of species)") + 
    theme_classic()
```

***

<br>

# Figure 17.5-4 Residual plots

***Using residual plots** to detect non-normality and unequal variance.*

<br>

## Blue tit cap color

The plot on the left is from a data set of blue tit cap color of offspring and parents. To begin, read and inspect the data.

```{r eval = TRUE}
capColor <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f5_4BlueTitCapColor.csv"), stringsAsFactors = FALSE)
head(capColor)
```

A quick residual plot can be obtained by plotting an ```lm()``` object with the ``plot()``` functon. 

```{r eval = TRUE}
capColorRegression <- lm(offspringColorScore ~ midparentColorScore, data = capColor)
plot(capColorRegression, which = 1)
```

Our Figure 17.5-4 can be reproduced using the following commands. First, fit the linear regression model using ```lm()``` and save the residuals in the data frame. Then plot the residuals against the explanatory variable using ```ggplot()```.

```{r eval = TRUE}
capColorRegression <- lm(offspringColorScore ~ midparentColorScore, data = capColor)
capColor$residuals <- residuals(capColorRegression)
ggplot(capColor, aes(midparentColorScore, residuals)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_hline(yintercept=0, color = "black") +
    labs(x = "Midparent color score", y = "Residuals") + 
    theme_classic()
```

<br>

## Cockroach neurons

Figure 17.5-4 (right) can be reproduced as follows. The data are firing rates of cockroach neurons at different temperatures. Begin by reading and inspecting the data.

```{r eval = TRUE}
roach <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f5_4CockroachNeurons.csv"), stringsAsFactors = FALSE)
head(roach)
```

Next, fit the linear regression model using ```lm()``` and save the residuals in the data frame. Then plot the residuals against the explanatory variable.

```{r eval = TRUE}
roachRegression <- lm(rate ~ temperature, data = roach)
roach$residuals <- residuals(roachRegression)
ggplot(roach, aes(temperature, residuals)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_hline(yintercept=0, color = "black") +
    labs(x = expression(Temperature~(degree*C)), y = "Residuals") + 
    theme_classic()
```

***

<br>

# Figure 17.6-3. Iris pollination

***Use a square root transformation** to improve the fit to assumptions of linear regression. The data are number of pollen grains received and floral tube length of an iris species*.

<br>

## Read and inspect data

```{r eval = TRUE}
iris <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f6_3IrisPollination.csv"))
head(iris)
```
<br>

## Residual plot untransformed data

A quick residual plot can be obtained by plotting an ```lm()``` object with the ``plot()``` command.. 

```{r eval = TRUE}
irisRegression <- lm(grainsDeposited ~ tubeLengthMm, data = iris)
plot(irisRegression, which = 1)
```

Our Figure 17.6-3 (left) can be reproduced with the following commands. We first fit the linear regression model using ```lm()``` and save the residuals in the data frame. Then we plot the residuals against the explanatory variable using ```ggplot()```.

```{r eval = TRUE}
irisRegression <- lm(grainsDeposited ~ tubeLengthMm, data = iris)
iris$residuals <- residuals(irisRegression)
ggplot(iris, aes(tubeLengthMm, residuals)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_hline(yintercept=0, color = "black") +
    labs(x = "Floral tube length (mm)", y = "Residuals") + 
    theme_classic()
```

<br>

## Square root transformation

We square-root transformed the single variable ```grainsDeposited``` and saved the resulting new variable (```sqRootGrains```) to the data frame. 

```{r eval = TRUE}
iris$sqRootGrains <- sqrt(iris$grainsDeposited + 1/2)
```

<br>

## Residual plot transformed data

To make Our Figure 17.6-3 (right) we fitted a linear regression to the tranformed y-variable with ```lm()``` and saved the residuals to the data fraom. Then we replotted the residuals. 

```{r eval = TRUE}
irisRegressionSqrt <- lm(sqRootGrains ~ tubeLengthMm, data = iris)
iris$residuals <- residuals(irisRegressionSqrt)
ggplot(iris, aes(tubeLengthMm, residuals)) + 
    geom_point(size = 3, col = "firebrick") + 
    geom_hline(yintercept=0, color = "black") +
    labs(x = "Floral tube length (mm)", y = "Residuals") + 
    theme_classic()
```

***

<br>

# Figure 17.8-1. Phytoplankton growth

***Fit a nonlinear regression curve having an asymptote** (Michaelis-Menten curve). The data are the relationship between population growth rate of a phytoplankton in culture and the concentration of iron in the medium*.

<br>

## Read and examine data

```{r eval = TRUE}
phytoplankton <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f8_1IronAndPhytoplanktonGrowth.csv"), stringsAsFactors = FALSE)
head(phytoplankton)
```

<br>

## Scatter plot

```{r eval = TRUE}
ggplot(phytoplankton, aes(ironConcentration, phytoGrowthRate)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = expression(paste("Iron concentration (", mu, "mol)")), 
         y = "Specific growth rate (no./day)") + 
    theme_classic()
```

<br>

## Fit a Michaelis-Menten curve

Use the ```nls``` (nonlinear least squares) to fit the curve. Provide a formula that also includes symbols (letters) for the parameters to be estimated. We use "a" and "b" to indicate the parameters of the Michaelis-Menten curve to estimate. The function includes an argument providing an initial guess of parameter values. The value of the initial guess is not so important -- here we choose ```a=1``` and ```b=1``` as initial guesses. The first function below carried out the model fit and save the results in an R object named ```phytoCurve```.

```{r eval = TRUE}
phytoCurve <- nls(phytoGrowthRate ~ a*ironConcentration / ( b+ironConcentration), 
	data = phytoplankton, list(a = 1, b = 1))
```

Use the ```summary``` command to view the parameter estimates, including standard errors and *t*-tests of null hypotheses that parameter values are zero.

```{r eval = TRUE}
summary(phytoCurve)
```

Many of the functions that can be used to extract results from a saved ```lm``` object work in the same way when applied to an ```nls``` object, such as ```predict```, ```residuals```, and ```coef```.

<br>

## Add curve to scatter plot

The function ```geom_smooth()``` has a nonlinear least squares method that allows you to input the same formula as in the ```nls()``` command above.

```{r eval = TRUE}
ggplot(phytoplankton, aes(ironConcentration, phytoGrowthRate)) + 
    geom_smooth(method = "nls", method.args = 
        list(formula = y ~ a * x / (b + x), start = list(a = 1, b = 1)), 
              data = phytoplankton, se = FALSE, col = "black") + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = expression(paste("Iron concentration (", mu, "mol)")), 
         y = "Specific growth rate (no./day)") + 
    theme_classic()
```

***

<br>

# Figure 17.8-2. Pond productivity

***Fit a quadratic curve** to the relationship between the number of plant species present in ponds and pond productivity*.

<br>

## Read and examine data.

```{r eval = TRUE}
pondProductivity <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f8_2PondPlantsAndProductivity.csv"), stringsAsFactors = FALSE)
head(pondProductivity)
```

<br>

## Scatter plot with quadratic curve

Provide ```geom_smooth()``` with the formula to add the curve to the scatter plot. Wrap the squared term in the formula with ```I()``` so that the arithmetic (```x^2```) is done before it is entered as a term in the model.

```{r eval = TRUE}
ggplot(pondProductivity, aes(productivity, species)) + 
    geom_smooth(method = "lm", formula = y ~ x + I(x^2), 
        color = "black", se = FALSE) +
    geom_point(size = 3, col = "firebrick") + 
    labs(x ="Productivity (g/15 days)", y = "Number of species") + 
    theme_classic()
```


<br>

## Fit quadratic function

Use ```lm()``` to fit a quadratic function to the data. Wrap the squared term with ```I()``` so that the arithmetic (```productivity^2```) is done before it is entered as a term in the model. The results of the model fit are saved in an ```lm()``` object ```productivityCurve```. Use ```summary()``` to extract estimated cowfficients and standard errors.

```{r eval = TRUE}
productivityCurve <- lm(species ~ productivity + I(productivity^2), 
	data = pondProductivity)
summary(productivityCurve)
```

***

<br>

# Example 17.8. Shrinking seals

**Fit a formula-free curve** (cubic spline) to the relationship between body length and age for female fur seals.

<br>

## Read and inspect data

```{r eval = TRUE}
shrink <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17e8ShrinkingSeals.csv"), stringsAsFactors = FALSE)
head(shrink)
```

<br>

## Scatter plot with spline

A small amount of vertical jitter was added to reduce overlap of points.

```{r eval = TRUE}
ggplot(shrink, aes(ageInDays, length)) + 
    geom_jitter(size = 0.3, col = "firebrick", height = 0.5) + 
    geom_spline(col = "black", df = 40) +
    labs(x = "Female age (days)", y = "Body length (cm)") + 
    theme_classic()
```

***

<br>

# Figure 17.9-1. Guppy cold death

***Fit a logistic regression** to the relationship between guppy mortality and duration of exposure to a temperature of 5 degrees C*.

<br>

## Read and inspect data

```{r eval = TRUE}
guppy <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter17/chap17f9_1GuppyColdDeath.csv"), stringsAsFactors = FALSE)
head(guppy)
```

<br>

## Frequency table

Tabulate mortality at different exposure times (Table 17.9-1).

```{r eval = TRUE}
table(guppy$mortality, guppy$exposureDurationMin, 
	dnn = c("Mortality","Exposure (min)"))
```

<br>

## Scatter plot logistic curve

The points are jittered sightly to minimize overlap.

```{r eval = TRUE}
ggplot(guppy, aes(exposureDurationMin, mortality)) + 
    geom_smooth(method = "glm", se = FALSE, color = "black",
        method.args = list(family = "binomial")) +
    geom_jitter(shape = 1, col = "firebrick", size = 3,
        height = 0.02, width = 0.5) + 
    labs(x = "Duration of exposure (min)", y = "Mortality") + 
    theme_classic()
```

<br>

## Fit a logistic regression

Use ```summary()``` to obtain estimates of regression coefficients with standard errors (Table 17.9-2).

```{r eval = TRUE}
guppyGlm <- glm(mortality ~ exposureDurationMin, data = guppy,
	family = binomial(link = logit))
summary(guppyGlm)
```

<br>

## 95% confidence intervals

Use ```confint()``` to calcaulate approximate 95% confidence intervals of parameters. 

```{r eval = TRUE, message=FALSE}
confint(guppyGlm)
```

<br>

## Predict mortality

Use ```predict()``` to predict probability of death (mortality) for a 10 min exposure duration, including a standard error of prediction.

```{r eval = TRUE}
predictMortality <- predict(guppyGlm, type = "response", se.fit = TRUE,
    newdata = data.frame(exposureDurationMin = 10))
data.frame(predictMortality)
```

<br>

## Estimate the LD50

The LD50 is the estimated dose at which half the individuals are predicted to die from exposure. The ```dose.p()``` function is from the ```MASS``` package. 

```{r eval = TRUE}
dose.p(guppyGlm, p = 0.50)
```

<br>

## Analysis of deviance table

The analysis of deviance table  (Table 17.9-3) is analogous to the ANOVA table. It includes a test of the null hypothesis of zero slope.

```{r eval = TRUE}
anova(guppyGlm, test = "Chi")
```

