---
title: "R code for examples in Chapter 2: <br>Correlation between numerical variables"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

*** 

We used R to analyze all examples in Chapter 16. We put the code here so that you can too.

Also see our [R lab on correlation](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Correlation_Regression.html).

<br>

# New methods on this page

<ul>
<li>Linear correlation:

```{r eval = FALSE}
cor.test(~ futureBehavior + nVisitsNestling, data = booby)
```

<li>Spearman rank correlation:

```{r eval = FALSE}
cor.test(~ impressivenessScore + years, data = trick, method = "spearman")
```
</ul>

***

# Load required packages 

We need to load the ```ggplot2``` package for graphs. We might need the ```dplyr``` package for subsetting a data set.

```{r message = FALSE}
library(ggplot2)
library(dplyr)
```

***

<br>

# Example 16.1. Flipping booby

***Estimate a linear correlation** between the number of non-parent adult visits experienced by boobies as chicks and the number of similar behaviors performed by the same birds when adult*.

<br>

## Read and inspect data

```{r eval = TRUE}
booby <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter16/chap16e1FlippingBird.csv"), stringsAsFactors = FALSE)
head(booby)
```

<br>

## Scatter plot

```{r eval = TRUE}
ggplot(booby, aes(nVisitsNestling, futureBehavior)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Events experienced while a nestling", y = "Future behavior") + 
    theme_classic()
```

<br>

## Correlation coefficient

The ```cor.test``` function computes a number of useful quantities, which we save in the object ```boobyCor```. The quantities can be extracted one at a time or shown all at once. We show the formula method here. 

```{r eval = TRUE}
boobyCor <- cor.test(~ futureBehavior + nVisitsNestling, data = booby)
boobyCor
```

If only the estimated correlation $r$ and its standard error are of interest, they can be obtained as follows. The calculation of standard error uses ```nrow(booby)``` to get the sample size for the correlation, but this will only be true if there are no missing values.

```{r eval = TRUE}
r <- boobyCor$estimate
r
SE <- sqrt( (1 - r^2)/(nrow(booby) - 3) )
unname(SE)
```

<br>

## Confidence interval

The 95% confidence interval for the correlation is included in the output of ```cor.test```. It can be extracted from the ```boobyCor``` calculated in an earlier step.

```{r eval = TRUE}
boobyCor$conf.int
```

***

<br>

# Example 16.2. Inbreeding wolves

***Test a linear correlation** between inbreeding coefficients of litters of mated wolf pairs and the number of pups surviving their first winter*.

<br>

## Read and inspect data

```{r eval = TRUE}
wolf <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter16/chap16e2InbreedingWolves.csv"), stringsAsFactors = FALSE)
head(wolf)
```

<br>

## Scatter plot

```{r eval = TRUE}
ggplot(wolf, aes(inbreedCoef, nPups)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Inbreeding coefficient", y = "Number of pups") + 
    theme_classic()
```

<br>

## Test of zero correlation

The results of the test are included in the output of ```cor.test```.

```{r eval = TRUE}
cor.test(~ nPups + inbreedCoef, data = wolf)
```

***

<br>

# Figure 16.4-1. Stream invertebrates

***Effect of the range of the data on the correlation coefficient** between population density of (log base 10 of number of individuals per square meter) and body mass (g) of different species of stream invertebrates*.

<br>

## Read and inspect data

```{r eval = TRUE}
streamInvert <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter16/chap16f4_1StreamInvertebrates.csv"), stringsAsFactors = FALSE)
head(streamInvert)
```

<br>

## Scatter plot

```{r eval = TRUE}
ggplot(streamInvert, aes(log10Mass, log10Density)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Log body mass", y = "Log population density") + 
    theme_classic()
```

<br>

## Effect of the range of the data

Here is the correlation coefficient for the full range of the data. The command uses ```cor.test``` but we extract just the correlation coefficient for this exercise.

```{r eval = TRUE}
cor.test(~ log10Density + log10Mass, data = streamInvert)$estimate
```

We want to obtain the correlation coefficient for the subset of the data corresponding to a log10Mass between 0 and 2. First subset the data we using either ```subset()``` in base R or ```filter()``` in the ```dplyr``` package.

```{r eval = TRUE}
streamInvertReduced <- subset(streamInvert, log10Mass >= 0 & log10Mass <= 2)
# or
streamInvertReduced <- filter(streamInvert, log10Mass >= 0, log10Mass <= 2)
```

Then recalculate the correlation coefficient.

```{r eval = TRUE}
cor.test(~ log10Density + log10Mass, data = streamInvertReduced)$estimate
```

***

<br>

# Example 16.5. Indian rope trick

***Spearman rank correlation** between impressiveness score of the Indian rope trick and the number of years elapsed bewteen the witnessing of the trick and the telling of it in writing*.

<br>

## Read and inspect data

```{r eval = TRUE}
trick <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter16/chap16e5IndianRopeTrick.csv"), stringsAsFactors = FALSE)
head(trick)
```

<br>

## Scatter plot

```{r eval = TRUE}
ggplot(trick, aes(years, impressivenessScore)) + 
    geom_point(size = 3, col = "firebrick") + 
    labs(x = "Years elapsed", y = "Impressiveness score") + 
    theme_classic()
```

<br>

## Spearman's rank correlation

The correlation is part of the output of the ```cor.test()``` function. See the next section for the result.

```{r eval = FALSE}
cor.test(~ years + impressivenessScore, data = trick, method = "spearman")
```

<br>

## Test of zero correlation

In this example, the variable ```impressivenessScore``` is a number score with lots of tied observations. Because of the ties, R will warn you that the *P*-value in the output is not exact.

```{r eval = TRUE}
cor.test(~ years + impressivenessScore, data = trick, method = "spearman")
```

