---
title: "R code for examples in Chapter 4: <br>Estimating with uncertainty"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

We used R to analyze all examples in Chapter 4. We put the code here so that you can too.

<br>
Also see our [lab on how to compute confidence intervals for means in R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Describing_data.html).

<br>

# New methods on this page

<ul>
<li>Take a random sample from a known population of measurements:

```{r eval = FALSE}
genesample100 <- sample(humanGeneLengths$size, size = 100,
    replace = FALSE)
```

<li>Standard error of the mean, assuming no missing values:

```{r eval = FALSE}
n <- length(genesample100)
sd(genesample100) / sqrt(n)
```

<li>Create a loop to repeatedly sample from a known population and calculate the mean each time:

```{r eval = FALSE}
for(i in 1:1000){
    temporarySample = sample(humanGeneLengths$size, 
        size = 100, replace = FALSE)
    mean(temporarySample)
    }
```

<li>Other new methods:
<ul>
<li>Add means and error bars to a strip chart.
</ul>
</ul>

***

<br>

# Load required packages 

We'll use the ```ggplot2``` add on package to make graphs. It needs to be loaded only once per session.

```{r message = FALSE}
library(ggplot2)
```

***

<br>

# Example 4.1. Lengths of human genes

*Describe the **parameters of a known population** of human gene lengths. Then, **take a random sample from the known population** to estimate the population mean.*

<br>

## Read and inspect data

Read the human gene length data, which we will use as our known population of measurements.

```{r eval = TRUE}
humanGeneLengths <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter04/chap04e1HumanGeneLengthsLongestTranscript.csv"), stringsAsFactors = FALSE)
head(humanGeneLengths)
```

<br>

## Draw histogram

We only want to plot the sizes of genes up to 15,000 nucleotides for this exercise. Use ```subset()``` for this purpose and put subset into a second data frame.

```{r eval = TRUE}
geneLengthsUpTo15K <- subset(humanGeneLengths, size <= 15000)
```

Histogram of the length of genes in the human genome (Figure 4.1-1). 

```{r eval = TRUE}
hist(geneLengthsUpTo15K$size, right = FALSE, col = "firebrick", 
     breaks = seq(0, 15000, 500), main = "", las = 1)
```

or

```{r eval = TRUE}
ggplot(data = geneLengthsUpTo15K, aes(x = size)) + 
    geom_histogram(fill = "firebrick", col = "black", 
        boundary = 0, closed = "left", binwidth = 500) + 
    labs(x = "Gene length (nucleotides)", y = "Frequency") + 
    theme_classic()
```

<br>

## Mean and standard deviation

These are the known population mean and standard deviations (Table 4.1-1). Just this once we must use the total number of genes ```N``` instead of ```N - 1``` in the denominator when calculating the variance and standard deviation, because we treat all the genes of the human genome as a population for this exercise, not as a sample. For this reason we can't use the built-in commands to calculate variance and standard deviation, because they divide by ```N - 1```.

```{r eval = TRUE}
meanGeneLength <- mean(humanGeneLengths$size)
meanGeneLength

N <- nrow(humanGeneLengths)
N

varGeneLength <- sum( (humanGeneLengths$size - meanGeneLength)^2 ) / N
sdGeneLength <- sqrt(varGeneLength)
sdGeneLength
```

Commands to put the mean and standard deviation into a table are shown here.

```{r eval = TRUE}
data.frame(Parameter = c("Mean", "Standard deviation"), 
	Value = c(meanGeneLength, sdGeneLength))
```

<br>

## Take a random sample of 100 genes

Take a single random sample from the population of genes. The argument ```replace = FALSE``` ensures that the same gene is not sampled twice. Save your random sample to a vector. Note: your sample won't be the identical to the one in the book, because each random sample is subject to sampling error.

```{r eval = TRUE}
geneSample100 <- sample(humanGeneLengths$size, size = 100, replace = FALSE)
```

<br>

## Histogram of sample

We plotted only genes up to 15,000 nucleotides in length.

The ```ggplot()``` command requires that the data be in a data frame, and so we use ```data.frame(geneSample100)``` in the command. The output might include a warning if the sample contained one or more genes larger than 15,000 nucleotides, which were left out. 

```{r eval = TRUE}
hist(geneSample100[geneSample100 <= 15000], right = FALSE, col = "firebrick", 
     breaks = seq(0, 15000, 500), main = "", las = 1)
```

or

```{r eval = TRUE}
ggplot(data = data.frame(geneSample100), aes(x = geneSample100)) + 
    geom_histogram(fill = "firebrick", col = "black", 
        boundary = 0, closed = "left", binwidth = 500) + 
    labs(x = "Gene length (nucleotides)", y = "Frequency") + 
    xlim(0,15000) +
    theme_classic()
```

<br>

## Sample mean and standard deviation 

Because of sampling error, your values won't be the identical to the ones in Table 4.1-2.

```{r eval = TRUE}
mean(geneSample100)

sd(geneSample100)
```

***

<br>

# Show the sampling distribution

*This shows how to take many random samples from the population to **visualize the sampling distribution** of the sample mean.*

<br>

## Create a loop 

Write a loop to take a large number of random samples of size $n = 100$ from the population. In each iteration, the sample mean is calculated and saved in a vector named ```results100``` (the samples themselves are not saved). The results vector is initialized before the loop. The term ```results100[i]``` refers to the ```i```th element of ```results100```, where ```i``` is a counter. This many iterations might take a few minutes to run on your computer.

```{r eval = TRUE}
results100 <- vector() 
for(i in 1:10000){
	temporarySample <- sample(humanGeneLengths$size, size = 100, 
		replace = FALSE)
	results100[i] <- mean(temporarySample)
	}
```

<br>

## Histogram of sample means

The commands give the relative frequencies, as in Figure 4.1-3. Your results won't be perfectly identical to ours in Figure 4.1-3 because 10,000 random samples is not a large enough number of iterations to obtain the true sampling distribution with extreme accuracy. 

```{r eval = TRUE}
ggplot(data = data.frame(results100), aes(x = results100)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 50) +
    labs(x = "Gene length (nucleotides)", y = "Probability") + 
    theme_classic()
```

<br>

## Compare different sample sizes

The following commands compare the sampling distribution of the mean for different sample sizes (Figure 4.1-4). The ```par()``` command formats the page

```{r eval = TRUE}
results20 <- vector()
for(i in 1:10000){
	  tmpSample <- sample(humanGeneLengths$size, size = 20, replace = FALSE)
	  results20[i] <- mean(tmpSample)
    }
ggplot(data = data.frame(results20), aes(x = results20)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 100) +
    labs(x = "Gene length (nucleotides)", y = "Probability") + 
    coord_cartesian(xlim = c(2000, 6000)) +
    theme_classic()

results100 <- vector()
for(i in 1:10000){
	  tmpSample <- sample(humanGeneLengths$size, size = 100, replace = FALSE)
	  results100[i] <- mean(tmpSample)
    }
ggplot(data = data.frame(results100), aes(x = results100)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 50) +
    labs(x = "Gene length (nucleotides)", y = "Probability") + 
    coord_cartesian(xlim = c(2000, 6000)) +
    theme_classic()

results500 <- vector()
for(i in 1:10000){
	  tmpSample <- sample(humanGeneLengths$size, size = 500, replace = FALSE)
	  results500[i] <- mean(tmpSample)
    }
ggplot(data = data.frame(results500), aes(x = results500)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 25) +
    labs(x = "Gene length (nucleotides)", y = "Probability") + 
    coord_cartesian(xlim = c(2000, 6000)) +
    theme_classic()

```

<br>

## Standard error of the mean

We can calculate the standard deviation of the sampling distributions (standard errors) based on the many rando samples of size 20, 100, and 500. Your results won't be perfectly identical to ours in Table 4.2-1 because 10,000 random samples is not a large enough number of iterations to obtain the true sampling distribution with extreme accuracy. 

```{r eval = TRUE}
data.frame("Sample size" = c(20,100,500), 
      "Standard error" = c(sd(results20), sd(results100), sd(results500)))
```

The standard error of gene length for the unique sample of 100 genes is as follows. The ```length``` command indicates the number of elements in a vector variable, which is the sample size if there are no missing (NA) elements in the vector. You won't get the same value for the standard error as we obtained because your unique random sample is not identical to ours.

```{r eval = TRUE}
n <- length(geneSample100)
se.mean <- sd(geneSample100) / sqrt(n)
se.mean
```

<br> 

## Approximate confidence intervals 

The commands below plot approximate 95% confidence intervals for the population mean based on 20 random samples of size $n = 100$ (Figure 4.3-1). For the population we used the genes whose sizes were less than or equal to 15,000 nucleotides. 

```{r eval = TRUE}
results <- data.frame(mean = rep(0,20), lower = rep(0,20), upper = rep(0,20))
for(i in 1:20){
	tmpSample <- sample(humanGeneLengths$size, size = 100, replace = FALSE)
	t <- t.test(tmpSample)
	results$mean[i] <- t$estimate
	results$lower[i] <- t$conf.int[1]
	results$upper[i] <- t$conf.int[2]
	}

plot(c(1:20) ~ mean, data = results, pch = 16, col = "red", yaxt = "n",
	xlim = c(min(results$lower), max(results$upper)), bty = "l", 
	xlab = "Gene length (number of nucleotides)", ylab = "")
lines(c(3511.457, 3511.457), c(0,20), lty = 2)
segments(results$lower, 1:20, results$upper, 1:20)
segments(results$lower, 1:20 - 0.25, results$lower, 1:20 + 0.25)
segments(results$upper, 1:20 - 0.25, results$upper, 1:20 + 0.25)

```

***

<br>

# Error bars Fig. 4.4-1

***Draw a strip chart with standard error bars** for the locust serotonin data. The data are from Chapter 2 (Figure 2.1-2).*

<br>

## Read and inspect the data.

```{r eval = TRUE}
locustData <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter02/chap02f1_2locustSerotonin.csv"), stringsAsFactors = FALSE)
head(locustData)
```

<br>

## Draw the strip chart

Use ```stat_summary()``` to overlay mean and standard error bars. We shifted the position of the bars using ```position_nudge()``` to eliminate overlap.

```{r eval = TRUE}
ggplot(locustData, aes(treatmentTime, serotoninLevel)) +
    geom_jitter(color = "firebrick", size = 3, width = 0.1) +
    stat_summary(fun.data = mean_se, geom = "errorbar", 
        width = 0.1, position=position_nudge(x = 0.2)) +
    stat_summary(fun.y = mean, geom = "point", 
        size = 3, position=position_nudge(x = 0.2)) +
    labs(x = "Treatment time (hours)", y = "Serotonin (pmoles)") + 
    theme_classic()
```


