---
title: "R code for example in Chapter 11: <br>Inference for a normal population"
author: 

#output: html_document
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height=4)
```

***

We used R to analyze all examples in Chapter 11. We put the code here so that you can too.

<br>
Also see our [lab on calculating the confidence interval for the mean using R](https://whitlockschluter3e.zoology.ubc.ca/RLabs/R_tutorial_Normal_and_sample_means.html).

<br>

# Load required packages 

We'll use the ```ggplot2``` add on package to make graphs.

```{r message = FALSE}
library(ggplot2)
```

***

<br>

# New methods on this page

<ul>
<li>One-sample $t$-test:

```{r eval = FALSE}
t.test(heat$temperature, mu = 98.6)
```

<li>Other new methods:
<ul>
<li>Confidence intervals for the variance and the standard deviation.
</ul>
</ul>

***

<br>

# Example 11.2. Stalk-eyed flies

***Confidence intervals for the population mean, variance, and standard deviation** using eye span measurements from a sample of stalk-eyed flies*.

<br>

## Read and inspect data

```{r eval = TRUE}
stalkie <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter11/chap11e2Stalkies.csv"), stringsAsFactors = FALSE)
stalkie
```

<br>

## Histogram

```{r eval = TRUE}
hist(stalkie$eyespan, right = FALSE, col = "firebrick", las = 1, 
	xlab = "Eye span (mm)", ylab = "Frequency", main = "")
```

or

```{r eval = TRUE}
ggplot(stalkie, aes(x = eyespan)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.2, 
        boundary = 0, closed = "left") + 
    labs(x = "Eye span (mm)", y = "Frequency") + 
    theme_classic()
```

<br>

## 95% CI for the mean

Appending ```$conf.int``` after the function ```t.test()``` causes R to give the 95% confidence interval for the mean.

```{r eval = TRUE}
t.test(stalkie$eyespan)$conf.int
```

<br>

## 99% CI for the mean

Adding the argument ```conf.level = 0.99``` changes the confidence level of the confidence interval.

```{r eval = TRUE}
t.test(stalkie$eyespan, conf.level = 0.99)$conf.int
```

<br>

## 95% CI for variance

Strangely, base R has no built-in function for the confidence interval of a variance, so must we compute it using the formula in the book:

```{r eval = TRUE}
df <- length(stalkie$eyespan) - 1
varStalkie <- var(stalkie$eyespan)
lower = varStalkie * df / qchisq(0.05/2, df, lower.tail = FALSE)
upper = varStalkie * df / qchisq(1 - 0.05/2, df, lower.tail = FALSE)
c(lower = lower, variance = varStalkie, upper = upper)
```

<br>

## 95% CI for std deviation

Take the square root of the confidence limits for the variance.

```{r eval = TRUE}
c(lower = sqrt(lower), std.dev = sqrt(varStalkie), upper = sqrt(upper))
```

***

<br>

# Example 11.3. Body temperature

*Uses a **one-sample *t*-test** to compare body temperature in a random sample of people with the "expected" temperature 98.6 °F*.

<br>

## Read and inspect data

```{r eval = TRUE}
heat <- read.csv(url("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter11/chap11e3Temperature.csv"), stringsAsFactors = FALSE)
head(heat)
```

<br>

## Histogram

```{r eval = TRUE}
hist(heat$temperature, right = FALSE, breaks = seq(97, 100.5, by = 0.5),
	col = "firebrick", las = 1, xlab = "Body temperature (°F)", 
	ylab = "Frequency", main = "")
```

or

```{r eval = TRUE}
ggplot(heat, aes(x = temperature)) + 
    geom_histogram(fill = "firebrick", col = "black", binwidth = 0.5, 
        boundary = 97, closed = "left") + 
    xlim(97, 100.5) +
    labs(x = "Body temperature (°F)", y = "Frequency") + 
    theme_classic()
```

<br>

## One-sample *t*-test

Calculate using the R command ```t.test()```. Use the argument ```mu = ``` to set the parameter value stated in the null hypothesis.

```{r eval = TRUE}
t.test(heat$temperature, mu = 98.6)
```
