---
title: "Displaying data: R code for Chapter 2 examples"
author:

<!--  output: html_document -->
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<code><!-- TEMPLATE --------------------------------------
# BOXES BEGIN
<div style="width:650px; padding:6px; border:2px solid black; bgcolor="#FFFFFF"">

<br>

### New methods on this page
Click on a function argument for a short description of its meaning. The variable names are plucked from the examples further below.

# Info with popups
<b>Topic</b>:
<code>command(arg1, arg2, arg3)</code>

<b>Other new methods</b>: 
etc</div>
# BOXES END

# BEGIN EXPERT POPUPS
To see how to produce the xxxxx (Figure xxx) click 
here

```{r eval = FALSE}
# Note.
commands...
```

# END EXPERT POPUPS
---------------------------------------------------- --></code>

Download the R code on this page as a single file <a href="../wp-content/rcode/chap02.r">here</a>.

<code><!-- ========================================================================================== --></code>
<div style="width:650px; padding:6px; border:2px solid black; bgcolor=" #ffffff""="">

<br>

### New methods on this page
Click on a function argument for a short description of its meaning. The variable names are plucked from the examples further below.

<b>Read data</b> from a comma-separated (.csv) file into a data frame. R will take columns having only numbers and make them numeric variables in the data frame. Columns having characters will be converted to factor variables, rather than to character variables.
<code>locustData <- read.csv(chap02f1_2locustSerotonin.csv)</code>

<b>Inspect data in a data frame</b>. The <code>head</code> function shows the first few lines of the data frame to confirm the data was read properly. The <code>nrow</code> function indicates the number of cases (rows) in the data frame.
<code>head(locustData)</code>
<code>nrow(locustData)</code>

<b>Basic strip chart</b> relating a numeric variable to a categorical variable:
<code>stripchart(serotoninLevel ~ treatmentTime, data = locustData, method = ,
&nbsp;&nbsp;&nbsp;&nbsp;vertical = TRUE)</code>

<b>Frequency table or contingency table</b>:

<b>Bar graph</b> for count data:
<code>barplot(tigerTable, ylab = , cex.names = 0.5)</code>

<b>Histogram</b> for a numeric variable:

<b>Scatter plot</b> showing association between two numeric variables:
<code>plot(sonAttractiveness ~ fatherOrnamentation, data = guppyFatherSonData)</code>

<b>Other new methods</b>:
Multiple-histogram plot.
Line plot.
Load an R package.
Set the desired ordering of categories of a factor variable in tables and graphs.

</div>
<code><!-- ========================================================================================== --></code>

<br>

### Figure 2.1-2. <a href="../wp-content/data/chapter02/chap02f1_2locustSerotonin.csv">Locust serotonin</a>
<em><b>Strip chart</b> of serotonin levels in the central nervous system of desert locusts that were experimentally crowded for 0 (the control group), 1, and 2 hours.</em>

<strong>Read the data and store in data frame</strong> (here named locustData). The following command uses <code>read.csv</code> to grab the data from a file on the internet (on the current web site).

```{r eval = FALSE}
locustData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f1_2locustSerotonin.csv"))
```

<strong>Show the first few lines of the data</strong>, to ensure it read correctly. Determine the number of cases in the data.

```{r eval = FALSE}
head(locustData)
nrow(locustData)
```

<strong>Draw a stripchart</strong> (the tilde "~" means that the first argument below is a formula, relating one variable to the other).

```{r eval = FALSE}
stripchart(serotoninLevel ~ treatmentTime, data = locustData, method = "jitter", 
	vertical = TRUE)
```

We show here

```{r eval = FALSE}
# Stripchart with options
par(bty = "l") # plot x and y axes only, not a complete box
stripchart(serotoninLevel ~ treatmentTime, data = locustData, vertical = TRUE, 
	method = "jitter", pch = 16, col = "firebrick", cex = 1.5, las = 1,
	ylab = "Serotonin (pmoles)", xlab = "Treatment time (hours)",
	ylim = c(0, max(locustData$serotoninLevel)))
# The following command calculates the means in each treatment group (time)
meanSerotonin = tapply(locustData$serotoninLevel, locustData$treatmentTime, mean)
# "segments" draws draws lines to indicate the means
segments(x0 = c(1,2,3) - 0.1, y0 = meanSerotonin, x1 = c(1,2,3) + 0.1, 
	y1 = meanSerotonin, lwd = 2)
```


***


<br>

### Figure 2.1-3. <a href="../wp-content/data/chapter02/chap02f1_3EducationSpending.csv">Education spending</a>
<em>A <b>bar graph</b> of education spending per student in different years in British Columbia.</em>

Read the data into a data frame named <code>educationSpending</code>, and inspect.

```{r eval = FALSE}
educationSpending <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f1_3EducationSpending.csv"))
head(educationSpending)
```

<strong>Draw a bar graph</strong>. The <code>names</code> argument generates numbers between 1998 and 2004 in 1-year increments, to be used as labels along the horizontal axis.

```{r eval = FALSE}
barplot(educationSpending$spendingPerStudent, names.arg = educationSpending$year,
	ylab = "Education spending ($ per student)")
```

A slightly fancier bar graph like that in Figure 2.1-3, which includes additional options, is shown here

```{r eval = FALSE}
# Bar graph with more options
barplot(educationSpending$spendingPerStudent, names.arg = educationSpending$year, 
	las = 1, col = "firebrick", ylim = c(0,8000),
	ylab = "Education spending ($ per student)")
```


***


<br>

### Example 2.2A. <a href="../wp-content/data/chapter02/chap02e2aDeathsFromTigers.csv">Deaths from tigers</a>
<em><b>Frequency table</b> and <b>bar graph</b> showing activities of 88 people at the time they were killed by tigers near Chitwan National Park, Nepal, from 1979 to 2006.</em>

Read the data into data frame named <code>tigerData</code>

```{r eval = FALSE}
tigerData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e2aDeathsFromTigers.csv"))
head(tigerData)
```

<strong>Generate a frequency table.</strong> The <code>sort</code> function is included to sort the categories by their frequencies.

```{r eval = FALSE}
tigerTable <- sort(table(tigerData$activity), decreasing = TRUE)
tigerTable
```

You can arrange the frequency table vertically.

```{r eval = FALSE}
data.frame(Frequency = tigerTable)
```

Use the <code>addmargins</code> command to include sums in your frequency table.

```{r eval = FALSE}
data.frame(Frequency = addmargins(tigerTable))
```

<strong>Draw a bar graph</strong>. The additional arguments <code>cex.names = 0.5</code> shrinks the axis labels by 50%, and <code>las = 2</code> flips the labels, so that they all fit in the window.

```{r eval = FALSE}
barplot(tigerTable, ylab = "Frequency", cex.names = 0.5, las = 2)
```

A slightly fancier bar graph of the data with more options, like that shown in Figure 2.2-1, is shown here

```{r eval = FALSE}
oldpar = par(no.readonly = TRUE) # stores a backup copy of current graph settings in "oldpar"
par(mar = c(8, 4, 4, 2) + 0.1)   # creates more room for labels below the x-axis
barplot(tigerTable, las = 2, col = "firebrick", cex.names = 0.8, ylim = c(0,50), 
	xlab = "", ylab = "Frequency (number of people)")
mtext("Activity", side = 1, line = 7, cex = 0.9) # adds the text under the x-axis
par(oldpar)  # reverts graph settings back to default
```


***


<br>

### Example 2.2B. <a href="../wp-content/data/chapter02/chap02e2bDesertBirdAbundance.csv">Desert bird abundance</a>
<em><b>Frequency table</b> and <b>histogram</b> illustrating the frequency distribution of bird species abundance at Organ Pipe Cactus National Monument. </em>

Read and inspect the data.

```{r eval = FALSE}
birdAbundanceData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e2bDesertBirdAbundance.csv"))
head(birdAbundanceData)
```

<strong>Generate a frequency table</strong> of the numeric bird abundance variable. The option <code>right = FALSE</code> ensures that abundance value 300 (for example) is counted in the 300-400 bin rather than in the 200-300 bin.

```{r eval = FALSE}
birdAbundanceTable <- table(cut(birdAbundanceData$abundance, 
	breaks = seq(0,650,by=50), right = FALSE))
birdAbundanceTable
```

The same table oriented vertically and including the sum.

```{r eval = FALSE}
data.frame(Frequency = addmargins(birdAbundanceTable))
```

<strong>Draw a histogram</strong> of bird abundances.

```{r eval = FALSE}
hist(birdAbundanceData$abundance, right = FALSE)
```

Commands to draw a histogram with more options, such as Figure 2.2-3, are here

```{r eval = FALSE}
# Histogram with options
hist(birdAbundanceData$abundance, right = FALSE, breaks = seq(0,650,by=50),	
	col = "firebrick", las = 1, xlab = "Abundance (No. individuals)",  
	ylab = "Frequency (No. species)", main = "")
```


***


<br>

### Figure 2.2-5. <a href="../wp-content/data/chapter02/chap02f2_5SalmonBodySize.csv">Salmon body size</a>
<em><b>Histograms</b> of body mass of 228 female sockeye salmon sampled from Pick Creek in Alaska. The same data are plotted for three different interval widths: 0.1 kg, 0.3 kg, and 0.5 kg.</em>

Read the data into data frame named <code>salmonSizeData</code>

```{r eval = FALSE}
salmonSizeData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f2_5SalmonBodySize.csv"))
head(salmonSizeData)
```

<strong>Histograms with different bin widths</strong> (0.1, 0.3, and 0.5).

```{r eval = FALSE}
hist(salmonSizeData$massKg, right = FALSE, breaks = seq(1,4,by=0.1), col = "firebrick")
hist(salmonSizeData$massKg, right = FALSE, breaks = seq(1,4,by=0.3), col = "firebrick")
hist(salmonSizeData$massKg, right = FALSE, breaks = seq(1,4,by=0.5), col = "firebrick")
```


***


<br>

### Example 2.3A. <a href="../wp-content/data/chapter02/chap02e3aBirdMalaria.csv">Bird malaria</a>
<em><b>Contingency table</b>, <b>grouped bar plot</b>, and <b>mosaic plot</b> showing the association between egg removal treatment and incidence of malaria in female great tits</em>.

Read the data from the data file and inspect.

```{r eval = FALSE}
birdMalariaData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e3aBirdMalaria.csv"))
head(birdMalariaData)
```

Optional step: <em>Set the desired order of treatment categories in tables and graphs</em>. The <code>factor</code> command allows us to order the levels so that the category "Egg removal" comes before "Control" in tables and graphs (categories are otherwise ordered alphabetically).

```{r eval = FALSE}
birdMalariaData$treatment <- factor(birdMalariaData$treatment, 
	levels= c("Egg removal", "Control"))
```

<strong>Create a contingency table</strong>. Use <code>table</code> with the names of two categorical variables as arguments, beginning with the response variable.

```{r eval = FALSE}
birdMalariaTable <- table(birdMalariaData$response, birdMalariaData$treatment)
birdMalariaTable
```

<strong>Include row and column sums</strong> in the contingency table.

```{r eval = FALSE}
addmargins(birdMalariaTable, margin = c(1,2), FUN = sum, quiet = TRUE)
```

<strong>Draw a grouped bar graph</strong> using the contingency table.

```{r eval = FALSE}
barplot(as.matrix(birdMalariaTable), beside = TRUE)
```

Commands to produce a grouped bar graph with more options are shown here

```{r eval = FALSE}
# Grouped bar graph with options
barplot(as.matrix(birdMalariaTable), beside = TRUE, space = c(0.1, 0.3),
	las = 1, xlab = "Treatment", ylab = "Frequency",
	col = c("firebrick", "goldenrod1"), legend.text = rownames(birdMalariaTable), 
	args.legend = list(x = 0.3, y = max(birdMalariaTable), xjust = 0))
```


<strong>Draw a mosaic plot</strong>. The "<code>t</code>" command flips (transposes) the table to ensure that the explanatory variable is along the horizontal axis.

```{r eval = FALSE}
mosaicplot( t(birdMalariaTable), col = c("firebrick", "goldenrod1"), 
	sub = "Treatment", ylab = "Relative frequency", 
	cex.axis = 1.1, main = "")
```


***


<br>

### Example 2.3B. <a href="../wp-content/data/chapter02/chap02e3bGuppyFatherSonAttractiveness.csv">Guppy father and son comparison</a>
<em><b>Scatter plot</b> of the relationship between the ornamentation of male guppies and the average attractiveness of their sons</em>.

Read the data from the file.

```{r eval = FALSE}
guppyFatherSonData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e3bGuppyFatherSonAttractiveness.csv"))
head(guppyFatherSonData)
```

<strong>Draw a scatter plot</strong> using the formula approach (with the "~").

```{r eval = FALSE}
plot(sonAttractiveness ~ fatherOrnamentation, data = guppyFatherSonData)
```

See here

```{r eval = FALSE}
# Scatter plot with options
plot(sonAttractiveness ~ fatherOrnamentation, data = guppyFatherSonData, 
	las = 1, pch = 16, col = "firebrick", cex = 1.5, bty = "l", 
	xlab = "Father's ornamentation", ylab = "Son's attractiveness")
```


***


<br>

### Example 2.3C. <a href="../wp-content/data/chapter02/chap02e3cHumanHemoglobinElevation.csv">Human hemoglobin and elevation</a>
<em><b>Strip chart</b>, <b>box plot</b>, and <b>multiple histograms</b> showing hemoglobin concentration in men living at high altitude in three different parts of the world and in a sea level population (USA)</em>.

Read the data.

```{r eval = FALSE}
hemoglobinData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e3cHumanHemoglobinElevation.csv"))
head(hemoglobinData)
```

Obtain the sample sizes in each of the four populations

```{r eval = FALSE}
table(hemoglobinData$population)
```

<b>Draw a strip chart</b> of the hemoglobin data.

```{r eval = FALSE}
stripchart(hemoglobin ~ population, data = hemoglobinData, method = "jitter",
    vertical = TRUE)
```

Commands for a fancier strip chart like that shown in Figure 2.3-4 are here

```{r eval = FALSE}
par(bty = "l")
stripchart(hemoglobin ~ population, data = hemoglobinData, 
	vertical = TRUE, method = "jitter", jitter = 0.2, pch = 1, col = "firebrick",
	las = 1, xlab = "Male population", ylab = "Hemoglobin concentration (g/dL)")
```


<b>Draw a box plot</b>, relating hemoglobin to population.

```{r eval = FALSE}
boxplot(hemoglobin ~ population, data = hemoglobinData)
```

A box plot with more options, like that in Figure 2.3-4, is shown here.

```{r eval = FALSE}
par(bty = "l")
boxplot(hemoglobin ~ population, data = hemoglobinData,
	col = "goldenrod1", boxwex = 0.5, whisklty = 1, outcol = "black", 
	outcex = 1, outlty = "blank", las = 1, 
	xlab="Male population", ylab = "Hemoglobin concentration (g/dL)")
```

Show the association between hemoglobin and population using the <strong>multiple histograms</strong> approach (Figure 2.3-5). The following commands use the <code>lattice</code> package, which must first be loaded.

```{r eval = FALSE}
library(lattice)
histogram(~ hemoglobin | population, data = hemoglobinData,
	layout = c(1,4), col = "firebrick", breaks = seq(10,26,by=1))
```

here

```{r eval = FALSE}
# Multiple histograms using base graphics, plotting them one at a time.
# The "oma" option adjusts the outer margins of the whole figure
# The "mar" option tweaks the margins within each of the 4 plots.
oldpar = par(no.readonly = TRUE) # make backup of default graph settings
par(mfrow = c(4,1), oma = c(4, 6, 2, 6), mar = c(2, 5, 4, 2))
hist(hemoglobinData$hemoglobin[hemoglobinData$population == "Andes"],
	col = "firebrick", las = 1, main = "Andes",  
	breaks = seq(10,26,by=1), ylab = "Frequency")
hist(hemoglobinData$hemoglobin[hemoglobinData$population == "Ethiopia"],
	col = "firebrick", las = 1, main = "Ethiopia",  
	breaks = seq(10,26,by=1), ylab = "Frequency")
hist(hemoglobinData$hemoglobin[hemoglobinData$population == "Tibet"],
	col = "firebrick", las = 1, main = "Tibet",  
	breaks = seq(10,26,by=1), ylab = "Frequency")
hist(hemoglobinData$hemoglobin[hemoglobinData$population == "USA"],
	col = "firebrick", las = 1, main = "USA",  
	breaks = seq(10,26,by=1), ylab = "Frequency")
mtext("Hemoglobin concentration (g/dL)", side = 1, outer = TRUE, padj = 1.5)
par(oldpar) # revert to default graph settings
```


***


<br>

### Example 2.4A. <a href="../wp-content/data/chapter02/chap02e4aMeaslesOutbreaks.csv">Measles outbreaks</a>
<em><b>Line graph</b> showing confirmed cases of measles in England and Wales from 1995 to 2011. Annual counts are quarterly.</em>

Read the data from file.

```{r eval = FALSE}
measlesData <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02e4aMeaslesOutbreaks.csv"))
head(measlesData)
```

<b>Drawing a line graph</b> uses the same command as a scatter plot but adds the <code>type = "l"</code> argument to draw a line graph instead.

```{r eval = FALSE}
plot(confirmedCases ~ yearByQuarter, data = measlesData, type="l") 
```

Commands to draw a fancier line plot (like that in Figure 2.4-1) are shown here

```{r eval = FALSE}
# Line plot with options
plot(confirmedCases ~ yearByQuarter, data = measlesData, 
	type = "l", las = 1, lwd = 1, bty = "l", xlab = "Year",
	ylab = "Number of cases", xaxp = c(1995,2012,2012-1995))
points(confirmedCases ~ yearByQuarter, data = measlesData, pch = 16, 
	col = "firebrick", cex = 0.7)
```

